---
title: "HMSC - Bird SMD model v1"
output: html_notebook
---

install Hmsc from CRAN 
```{r}
if(!"Hmsc" %in% installed.packages()){
  ans <- readline(prompt = "Hmsc is not installed. Do you want to install it now (y/n)")
  if(ans == "y"){
   install.packages("Hmsc") 
  }
  }
```


```{r}
library(Hmsc)
library(sf)
library(tidyr)
library(dplyr)
library(corrplot)
library(here)
library(ggplot2)
library(readr)
library(raster)
library(stringr)
library(job)

source(here("scripts", "helper_functions.R"))
```

#load data

## Species table
+ yearly abundances for the linecounts for all of Sweden (2012-2018)
+ exported by `Species_matrix.Rmd`

we have excluded all Birds that are observed (across the study region)
+ with < 5 individuals per year 
+ less than 5% of routes per year

We have also excluded a range of birds that are not associated with the habitat in our study (see `Species_matrix.Rmd` for details)

The matrix contains 96 Species. 

```{r}
# SpeciesMatrix
#SpeciesMatrix <- read_tsv(here("data", "Bird_spec_matrix_points_filtered_Sweden.tsv"))
SpeciesMatrix <- read_tsv(here("data", "Bird_spec_matrix_lines_filtered_Sweden.tsv"))

#FBI classification
FBI_meta <- read_tsv(here("data", "FBI_meta.txt"))
```


## Land-use data 

+ Landuse types (number of 10m^2 pixel) for 500m buffer around each point for all routes in Sweden. 
+ exported by `Join_LandUsed.Rmd`
+ Based on NMD (`NMD_characterization.Rmd`) and Application Data (`SJB_characterization.Rmd`)
+ 2012-2018 but only agricultural land changes, rest is constant

LandUse from NMD and SJB have been grouped into `r ncol(LandUse) - 3` LandUse classes:

**Forest (from NMD)**

"Ädellövskog"
"Barrblandskog"
"Granskog"
"Lövblandad_barrskog"
"Triviallövskog"
"Triviallövskog_med_ädellövinslag"
"Tallskog"
"Temporärt_ej_skog"

**Other Landtype (NMD)**

"Öppen_våtmark"
"Hav"     
"Exploaterad_mark"            
"Sjö_och_vattendrag"
"Övrig_öppen_mark_med_vegetation"
"Övrig_öppen_mark_utan_vegetation"

**Agriculture**

"Fallow"                                                    
"Ley"
"Pasture"
"Pasture_with_Trees"
"Spring_cereal"
"Other"                            
"Tuber_Root"
"Winter_cereal"  
 

```{r}
#LandUse data OBS match to loaded species matrix!!!
LandUse <-  read_csv(here("data", "LandUse_joint_lines.csv"))
#LandUse <-  read_csv(here("data", "LandUse_joint_points.csv"))

JKP_karta <- readRDS(here("data", "JKP_karta.RDS"))


```


##Climate
Climate data for all routes, taken from WorldClim. See script `Worldclim_characterization.Rmd for details`

```{r}
#Climate
WorldClim <- read_tsv(here("data", "WorldClimRoutes.csv")) %>% 
  dplyr::select(karta, WCS_5, WCS_12)

# here I select 
# WCS_5 (the maximum temperature of the warmest month) and 
# WCS_12 (average annual precipitation)
# following Regos et al 2019 (Scientific Reports) --> just for testing

#Route Coordinates
Standard_geo <- read_sf(here("data", "Standard_routes_geo.gpkg")) %>% 
  #filter(type == "point") %>% 
  filter(type == "line") %>% 
  dplyr::select(-type)

# st_distance() for distance matrix in m
# st_coordinates() to extract absolute latitude
```


# HMSC Model for Jönköping

We follow the example of: 

Tikhonov, Gleb, Øystein H Opedal, Nerea Abrego, and Aleksi Lehikoinen. “Joint Species Distribution Modelling with the R‐package Hmsc,” Methods in Ecology and Evolution

(Appendix 2)

## Model using species abundance data

### Species matrix
Y = site x species matrix 

We keep only the routes in the selected region. 

```{r}

Y = SpeciesMatrix 

#order Y by year, point, route 
Y <- arrange(Y, yr, karta, position)

#collate the variables to make unique row identifiers
rowID_df <- Y %>% dplyr::select(1:3)
rowID <- 
  rowID_df %>% 
  mutate(rowID = paste(yr, karta, position, sep = "_")) %>% 
  pull(rowID)

#transform Y to dataframe with rownames
Y <- as.matrix(Y[,-c(1:3)])
rownames(Y) <- rowID

#list of species
colnames(Y)

```

###Environmental co-variates

XData is the dataframe of environmental co-variates. 

We include the Landuse types as continuous environmental co-variates 
We also include two climatic variables

```{r}

XData =
LandUse %>% 
  left_join(WorldClim) %>% #join LandUse and Climate Data
  mutate(year = as.numeric(gsub("SJB_(\\d{4})", "\\1", year))) %>%  #extract numeric year 
  mutate(rowID = paste(year, karta, position, sep = "_")) %>% 
  filter(rowID %in% rownames(Y)) %>% #subset to points in Species matrix
  arrange(year, karta, position) %>% 
  dplyr::select(-karta, -position, -year) 

all(sort(XData$rowID) == sort(rownames(Y)))
```

```{r}
XData[,colnames(XData) != "rowID"] %>% 
  cor %>% 
  corrplot("ellipse", type = "lower", diag = F) #addCoef.col = "black", number.digits = 2)

XData %>% 
  pivot_longer(-"rowID") %>% 
  ggplot(aes(x = value))+
  geom_histogram()+
  facet_wrap(~name, scale = "free_x")
```


### simplified model

summarise LandUse to fewer categories

```{r}
# XData <- 
# XData %>% 
#   mutate(Lövskog = 
#            Ädellövskog +
#            Triviallövskog +
#            Triviallövskog_med_ädellövinslag,
#          Barrskog = 
#            Lövblandad_barrskog+
#            Tallskog+
#            Granskog+
#            Barrblandskog,
#          Cereal = 
#            Spring_cereal+
#            Winter_cereal+
#            Tuber_Root,
#          Open = 
#            Fallow+
#            Öppen_våtmark+
#            Övrig_öppen_mark_med_vegetation+
#            Övrig_öppen_mark_utan_vegetation,
#          Clearcut = 
#            Temporärt_ej_skog
#            ) %>% 
#   dplyr::select(rowID, Lövskog, Barrskog, Clearcut, Pasture, Pasture_w_Trees, Ley, Cereal, Open, WCS_5, WCS_12)
  
```


we do not include the following variables in the model: 

+ Hav
+ Övrig_öppen_mark_utan_vegetation
+ Sjö_och_vattendrag
+ Tuber_Root
+ Other


## transform XData

```{r}
hab_vars <- colnames(XData)[1:22]
hab_vars <- hab_vars[!hab_vars %in% c("Hav", "Övrig_öppen_mark_utan_vegetation",
                                     "Sjö_och_vattendrag", "Tuber_Root", "Other")]

XData <- dplyr::select(XData, hab_vars, starts_with("WCS"), "rowID")

### for point buffer ###
#Ncells <- pi*500^2/100 # number of pixels in each buffer

### for line buffer ###
Ncells <-  1000*500/100 # number of pixels in each buffer

#standardise habitats in model to fraction of area
XData <- 
XData %>% 
  mutate(across(one_of(!!hab_vars), ~.x/Ncells))

XData %>% 
  mutate(rs = rowSums(.[,hab_vars])) %>% 
  ggplot(aes(x = rs))+
  geom_histogram()+
  geom_vline(xintercept = 0.8)

#exclude lines were habitats in model together make less than 60% of area
line_exclude <- 
  XData %>% 
  filter(rowSums(.[,hab_vars]) < 0.8) %>% 
  pull(rowID)


length(line_exclude) / nrow(XData) * 100

#exclude from XData
XData <- XData %>% 
  filter(!rowID %in% line_exclude)

#exclude from Y
Y <- Y[!rownames(Y) %in% line_exclude,]

```

show distribution

```{r}
XData %>% 
  pivot_longer(-rowID) %>% 
  ggplot(aes(x = value))+
  geom_histogram()+
  facet_wrap(~name, scales = "free")

ggsave(here("plots", "var_dist_buffers.jpeg"))
```

 
 


## check for colinearity

No strong co-linearities among the environmental variables

```{r}
#jpeg(here("plots", "var_cor.jpg"))
cor(XData[,c(hab_vars, "WCS_5", "WCS_12")]) %>% 
  corrplot("ellipse", type = "lower", diag = F, addCoef.col = "black", number.digits = 2)
#dev.off
```

###Study Design

We have 3 random factors

+ year (temporal explicit)
+ route (spatial explicit at route level)
+ sampling line (factor. The exact transect on which the birds have been observed. Allows us to estimate residual association between species [potential biotic interactions])


We could include another (non-spatial) random level at the level of karta, but I am not sure if that would make any sense. 


```{r}

studyDesign = rowID_df[3:1] #the order is position within karta within year

#combine position and karate to get unique sampling location
studyDesign <- 
studyDesign %>% 
  mutate(karta2 = karta) %>% 
  unite(sampling_loc, karta2, position) %>% 
  mutate(yr = paste("year_", yr, sep = "")) %>% 
  mutate_all(as.factor) %>% 
  as.data.frame() 

rownames(studyDesign) <- paste(gsub("year_(\\d+)", "\\1", studyDesign$yr), studyDesign$sampling_loc, sep = "_")

studyDesign <- studyDesign[row.names(studyDesign) %in% row.names(Y), ] %>% droplevels()
```



###Random effect structure

We include the `sampling_loc` as sample unit random factor
We include `karta` as spatial explicit random factor 
We include `yr` as temporal explicit random factor

```{r}
#spatial random level for routes at the route level
xycoords = 
  Standard_geo %>% 
  filter(karta %in% levels(studyDesign$karta)) %>% 
  group_by(karta) %>% 
  summarise(geom = st_combine(geom)) %>% 
  st_centroid() 
  
xycoords_M <- st_coordinates(xycoords)
rownames(xycoords_M) <- xycoords$karta

#spatial Random level for routes
rL1 = HmscRandomLevel(sData=xycoords_M)

#rL1$nfMin = 5
#rL1$nfMax = 10

#temporal random level
t_coords = matrix(unique(rowID_df$yr), ncol = 1)
row.names(t_coords) <- unique(studyDesign$yr)

rL2 = HmscRandomLevel(sData=t_coords)
  
#unstructured Random level of points 
rL3 = HmscRandomLevel(units = unique(studyDesign$sampling_loc))
```

###Formula for Env data

linear effect of all habitat variable and the two climatic variables + a quadratic effect of Temperature and precipitation

```{r}

XFormula = ~
  #Forest
  Ädellövskog + Barrblandskog + Granskog + Lövblandad_barrskog + Tallskog + Triviallövskog +
  Temporärt_ej_skog + Triviallövskog_med_ädellövinslag +
  #other landuse in NMD
  Exploaterad_mark + Öppen_våtmark + Övrig_öppen_mark_med_vegetation +
  #agricultural landuse
  Fallow + Ley + Pasture + Pasture_w_Trees + Spring_cereal + Winter_cereal +
  #Climate
  WCS_5 + WCS_12 

# XFormula = ~
#   #Forest
#   Lövskog + Barrskog + Clearcut +
#   #other landuse in NMD
#   Open + #Other +
#   #agricultural landuse
#   Pasture + Ley + Cereal + 
#   #Climate
#   WCS_5 + WCS_12
```


###Formula for Trait data

```{r}
#TrFormula = ~Migration + logBM + Wing_to_Length_ratio
```

###construct Models

m1 = full model pa
```{r}

Y_pa <- Y
Y_pa[Y_pa > 0] <- 1 

m1 = Hmsc(Y = Y_pa,
          XData = data.frame(XData[, colnames(XData) != "rowID"]), 
          XFormula = XFormula,
          distr = "probit",
          studyDesign = studyDesign,
          ranLevels = list(karta = rL1, yr = rL2, sampling_loc = rL3))

```

m2 = full model conditional of presence

```{r}
#rL1$nfMax = 4

Y_abund <- Y
Y_abund[Y_abund == 0] <- NA
Y_abund <- log(Y_abund)

m2 = Hmsc(Y = Y_abund,
          XData = data.frame(XData[, colnames(XData) != "rowID"]), 
          XFormula = XFormula,
          distr = "normal",
          studyDesign = studyDesign,
          ranLevels = list(karta = rL1, yr = rL2, sampling_loc = rL3))
```

export models
```{r}
full_model <- list(m1 = m1, m2 = m2)
save(full_model, file = here("model", "Full_models.RData"))
```


##run model 

I run the model on the cluster
```{r, eval = FALSE}
if(!dir.exists(here("model"))) dir.create(here("model"))

load(here("model", "Full_models.RData"))

base_path <- here("model", "")

job::job(HMSC_out = {
  
thin_ = c(100)
samples = 100
nChains = 10
  
set.seed(3450)
ptm = proc.time()

for(n in thin_){
  
  for (i in seq_along(full_model)){
    
  if(!TRUE %in% grepl(paste(names(full_model)[i],n,"", sep = "_"), list.files(base_path))){
  
  m = full_model[[i]]
  mod = sampleMcmc(m, samples = samples, thin = n,
                  transient = ceiling(0.5*samples*n),
                  nChains = nChains, nParallel = nChains,
                  updater=list(GammaEta = FALSE),
                  verbose = 5)

  computational.time = proc.time() - ptm

  save(mod, file=paste(base_path, names(full_model)[i], "_", n, "_", round(computational.time[["elapsed"]]), sep = ""))
    
  }

  }
}
})

```

## load models

load models
```{r}
model_files <- list.files(here("model"))
model_files <- model_files[grepl("m[1,2]_100_\\d+", model_files)]
model_files <- model_files[c(1,3)]

models <- 
lapply(model_files, function(x){
  filename<-gsub("(.+)_sam.+", "\\1", x)
  wd<-here("model", x)
  load(wd)
  assign(filename, mod)
})

names(models) <- gsub("(m\\d).+", "\\1", model_files)

```

#model evaluation

convert to coda object

```{r}
models_post <- lapply(models, convertToCodaObject)
```


```{r}
names(models_post$m1)
```

the fitted model has parameter estimates for the following parameters: 

Beta: species niches
Gamma: influence of traits on niches
V: residual covariance of species niches
Sigma: residual variance
Eta: Site loadings
Lambda: Species loadings
Omega: species associations
Alpha: spatial scale of site loadings
?? Psi: ?? Local shrinkage of species loadings ??
Delta: Global shrinkage of species loadings


## convergence


Here we visualize the effective sample size and gelman diagnostic for the beta parameters. We sampled 10 chains and recorded 200 samples from each chain (with thinning) so the maximum sample size is 2000. The effective sample size tells us how many 'independent' samples from the posterior distribution we have (after accounting for autocorrelation of the chains) and helps us estimate whether we can trust the mean posterior or not. 

The effective sample size can be higher than the number of samples if the samples are negatively correlated

The Gelman diagnostic (telling us how well the different chains mixed) should be close to 1.

```{r}

es.beta = lapply(models_post, function(x) effectiveSize(x$Beta))
ge.beta = lapply(models_post, function(x) gelman.diag(x$Beta,multivariate=FALSE)$psrf) 
# es.omega = lapply(models_post, function(x) effectiveSize(x$Omega[[1]]))
# ge.omega = lapply(models_post, function(x) gelman.diag(x$Omega[[1]],multivariate=FALSE)$psrf)

```


effective Size of the Beta parameters
```{r}
es.beta_df <- 
  es.beta %>% 
  bind_rows(.id = "model") %>%
  pivot_longer(-model, names_to = "parameter", values_to = "es") %>% 
  mutate(bird = gsub(".+\\),(.+)\\(.+", "\\1", parameter)) %>%
  mutate(bird = str_trim(bird)) %>% 
  mutate(fixed_effect = gsub("B\\[\\(?(\\w+)\\)? \\(.+", "\\1", parameter)) %>% 
  mutate(fixed_effect = gsub(".+(WCS_\\d{1,2}), (degree = \\d).+", "\\1_\\2",  fixed_effect))

es.beta_df %>% 
  ggplot(aes(x = fixed_effect, y = es, fill = model))+
  geom_violin(draw_quantiles = c(0.05, 0.5, 0.95))+
  geom_hline(yintercept = 1000)+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=0.5))+
  labs(title = "effective sample size (max = 2000) - Beta")+
  scale_fill_brewer(palette = "Set1")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
ge.beta_df <- 
ge.beta %>% 
  lapply(., function(x){
    as_tibble(x, rownames = "parameter")
  }) %>% 
  bind_rows(.id= "model") %>%
  pivot_longer(cols = one_of("Point est.", "Upper C.I."), names_to = "est", values_to = "val") %>% 
  mutate(bird = gsub(".+\\),(.+)\\(.+", "\\1", parameter)) %>%
  mutate(bird = str_trim(bird)) %>% 
  mutate(fixed_effect = gsub("B\\[\\(?(\\w+)\\)? \\(.+", "\\1", parameter)) %>% 
  mutate(fixed_effect = gsub(".+(WCS_\\d{1,2}), (degree = \\d).+", "\\1_\\2",  fixed_effect))


ggplot(ge.beta_df, aes(x = fixed_effect, y = val, fill = model))+
  geom_violin()+
  facet_wrap(~est, ncol = 1)+
  labs(title = "Gelman diagnostic (min = 1) - Beta")+
  scale_fill_brewer(palette = "Set1")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  

```



```{r}
es.alpha = lapply(models_post, function(x) effectiveSize(x$Alpha[[1]]))

es.alpha %>% 
  bind_rows(, .id = "model") %>% 
  #separate(col = model, into = c("model", "thin"), sep = "_thin") %>% 
  pivot_longer(cols = starts_with("Alpha1"), names_to = "Alpha", values_to = "val") %>%
  ggplot(aes(x = Alpha, y = val, colour = model))+
  geom_point()+
  #facet_wrap(~model, scales = "free")+
  scale_colour_brewer(palette = "Set1")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=0.5))
  

```


## model fit

The metrics of model fit that are computed depend on the type of response variable (count vs presence / absence). We modelled counts and presence absences seperately (hurdle model). As such HMSC gives us the following measures of model fit:

$RMSE gives the root mean square error between the predicted and observed values (median posterior abundance estimate compared to observed abundance across all sampling units, by species). The units are in the units of the predicted value (counts of birds at a single point count)

$SR2 is pseudo-R2 computed as the squared spearman correlation between observed and predicted values, times the sign of the correlation. 

$0.X are metrics that are calculated to predict the presence / absence (predicted and observed counts are truncated to p/a so the fit relates to how well the model can predict where the species can be found)

  $0.AUC measures the area under the curve. 
  $0.TjurR2 is the Tjur R square. 
  
Both are measures of discrimination that measure how well occurrence probabilities can discriminate between occupied or empty sampling units. If the model prediction is not better than by chance, 0.AUC will approach 0.5 while the TjurR2 will approach 0. A perfect model will approach 1 for both AUC and Tjur R2. 

$0.RSME is the root mean square error calculated on presence/absence area. The unit changes as now the models n predicts p/a. RSME thus gives the fraction of sites that are miss-predicted (*** I think ***)

$C.X is the opposite. It measures the predictive performance of the model for the count data on sites where the species is present. 

$C.SR2 thus gives the pseudo-R2 excluding all sites from which the species is absent

$C.RMSE gives the average discrepancy between predicted abundance and modelled abundance for sites were the species is present. 

```{r}

preds = lapply(models, function(x) {computePredictedValues(x,expected = TRUE)})

MF <- lapply(seq_along(models), function(x) evaluateModelFit(models[[x]], preds[[x]]))

MF_df <- 
lapply(seq_along(MF), function(x) {
  as_tibble(MF[[x]]) %>% 
    mutate(model = names(models)[x]) %>% 
    mutate(Species = colnames(preds[[1]]))
}) %>% 
  bind_rows() %>% 
  mutate(across(where(is.numeric), ~signif(.x, 2))) %>% 
  relocate(Species) %>% 
  pivot_longer(cols = !one_of("Species","model"), names_to = "metric", values_to = "val") 


ggplot(MF_df, aes(x = metric, y = val, colour = model))+
  geom_point(position = position_jitter(width = 0.1), alpha = 0.7)+
  #facet_wrap(~metric, scales = "free_y")+
  theme_bw()+
  scale_color_brewer(palette = "Set1")

```

## cross validation

```{r}
fit <- readRDS(here("model", "cross_val_models_full.rds"))

fit_cv <- 
as_tibble(fit) %>% 
  mutate(Species = colnames(preds[[1]])) %>% 
  mutate(across(where(is.numeric), ~signif(.x, 2))) %>% 
  pivot_longer(cols = !one_of("Species"), names_to = "metric", values_to = "val") %>% 
  left_join(MF_df, by = c("Species", "metric"), suffix = c("_cv", "_pred")) %>% 
  dplyr::select(-model) %>% 
  pivot_longer(starts_with("val")) 

p1 <- 
ggplot(fit_cv, aes(x = name, y = value))+
  geom_line(position = position_jitter(width = 0.05), 
            aes(group = Species), colour = "grey", size = 0.2)+
  geom_point(position = position_jitter(width = 0.05), alpha = 0.7, colour = "dodgerblue4")+
  facet_wrap(~metric)+
  theme_bw()

ggsave(here("model", "figures", "fit.jpg"), p1, height = 4, width = 6)
 
p1

```

```{r}

spec_fit <- 
  fit_cv %>% 
  filter(metric == "AUC", name == "val_cv") %>% 
  arrange(value) %>% 
  pull(Species)

p2 <- 
fit_cv %>% 
  mutate(Species = factor(Species, levels = spec_fit)) %>%
  ggplot(aes(x = Species, y = value, colour = name))+
  geom_point()+
  coord_flip()+
  facet_wrap(~metric)+
  scale_color_brewer(palette = "Set1")+
  theme_bw()
  
ggsave(here("model", "figures", "fit_by_spec.jpg"), p2, height = 10, width = 8)

p2
```


## Variance partitionig


```{r}

VP <- 
lapply(models, computeVariancePartitioning, 
       group = c(rep(1,17), rep(2,2)), 
       groupnames = c("Habitat", "Climate")) 

VP <- 
lapply(VP, function(x) as_tibble(x$vals, rownames = "factor")) %>% 
  bind_rows(.id = "model")

VP_p2 <- 
VP %>% 
  pivot_longer(!one_of("factor", "model", "thin")) 

bird_lev <- 
VP_p2 %>% 
  left_join(rename(FBI_meta, name = Englishname)) %>% 
  filter(factor == "Habitat") %>%
  filter(model == "m1") %>% 
  arrange(Europe, value) %>% 
  pull(name) 


p2 <- 
VP_p2 %>% 
  mutate(factor = factor(factor, 
                         levels = c("Climate", "Random: sampling_loc",
                                            "Random: yr","Random: karta","Habitat"))) %>% 
  left_join(rename(FBI_meta, name = Englishname)) %>% 
  mutate(name = factor(name, levels = bird_lev)) %>% 
  ggplot(aes(x = name, y = value, fill = factor))+
  geom_bar(stat = "identity", position = "stack")+
  scale_fill_brewer(palette = "Set1")+
  facet_grid(model*Europe~., scales = "free",  space = "free_y")+
  theme_bw()+
  coord_flip()+
  theme(legend.position = "top")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=0))


ggsave(here("model", "figures", "varpar_stacked.jpg"), p2, height = 16, width = 7)

p2
```

## residual species association

```{r}
Omega_cor_list <- lapply(models, computeAssociations)

Omega_cor_list_95 <- lapply(Omega_cor_list, function(x){
  {(x[[1]]$support > 0.95) * x[[1]]$mean}
})

mod_names = names(Omega_cor_list_95)

#jpeg(here("model", "figures", "resi_corr_stacked.jpg"))

lapply(seq_along(Omega_cor_list_95), function(x) corrplot(Omega_cor_list_95[[x]], tl.cex = 0.5, tl.col = "black", title = mod_names[x], mar=c(0,0,2,0))) %>% invisible()

#dev.off()

```

## parameter estimates

We have a look at the parameter estimates for the beta parameters

```{r}

fixed <-tibble(name = c(
"Ädellövskog", "Triviallövskog_med_ädellövinslag", "Triviallövskog",
"Lövblandad_barrskog", "Barrblandskog", "Tallskog","Granskog",
"Temporärt_ej_skog","Övrig_öppen_mark_med_vegetation","Öppen_våtmark",
"Exploaterad_mark","Fallow","Ley","Spring_cereal","Winter_cereal",
"Pasture","Pasture_w_Trees",
"WCS_12","WCS_5"),
short = c("Ädel.skog", "Triv_med_ädel", "Trivial.skog",
"Löv_med_barr", "Barrbland", "Tall","Gran",
"Clearcut","Övrig_öppen","Öppen_våtmark",
"Exploat._mark","Fallow","Ley","Spring_cereal","Winter_cereal",
"Pasture","Past_w_Trees",
"WCS_12","WCS_5"))

#function to get quantiles
###
quibble2 <- function(x, q = c(0.1, 0.5, 0.9), na.rm = TRUE) {
  tibble("{{ x }}" := quantile(x, q, na.rm), "{{ x }}_q" := q)
}
###

Beta_est <- 
lapply(models_post, function(x){
  lapply(x$Beta, function(y) {
    as.data.frame(y)}) %>% 
      bind_rows(.id = "chains")}) %>% 
  bind_rows(.id = "model") %>% 
  pivot_longer(cols=starts_with("B"), names_to = "parameter") 


Beta_est_sum <- 
  Beta_est %>% 
  group_by(model, parameter) %>% 
  summarise(quibble2(value)) %>% 
  pivot_wider(names_from = value_q, values_from = value)

Beta_est_sum <- 
Beta_est_sum %>% 
  mutate(bird = gsub(".+\\),(.+)\\(.+", "\\1", parameter)) %>%
  mutate(bird = str_trim(bird)) %>% 
  mutate(fixed_effect = gsub("B\\[\\(?(\\w+)\\)? \\(.+", "\\1", parameter)) %>% 
  filter(fixed_effect != "Intercept") %>% 
  mutate(fixed_effect = factor(fixed_effect, 
                               levels = fixed$name, 
                               label = fixed$short))

p3 <-  
  Beta_est_sum %>%
  left_join(rename(FBI_meta, bird = Englishname)) %>% 
  mutate(sig = case_when(`0.5` < 0 & `0.9` < 0 ~ "negative",
                           `0.5` > 0 & `0.1` > 0 ~ "positive",
                           TRUE ~ "non. sig.")) %>% 
  mutate(sig = factor(sig, levels = c("positive", "negative", "non. sig."))) %>% 
  ggplot(aes(y = bird, x = `0.5`, xmin = `0.1`, xmax = `0.9`, colour = sig))+
  geom_point(position = position_dodge(width = 0.8), size = 0.5)+
  geom_errorbarh(position = position_dodge(width = 0.8), size = 0.5)+
  geom_vline(xintercept = 0, size = 0.2, linetype = "dashed")+
  facet_grid(Europe~fixed_effect*model, scales = "free", space = "free_y")+
  scale_color_manual(values =  c("yellowgreen", "tomato3", "grey"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90))
  
  
ggsave(here("model", "figures", "beta_estimates.jpg"), p3, height = 10, width = 30)

Beta_est_sum %>% 
  write_tsv(here("data", "betapar,txt"))

p3


```




```{r}

jpeg(here("model", "figures", "corr_mat_beta_estimates_hm.jpg"), width = 1000, height = 1000)

Beta_est_sum %>% 
  filter(model == "m2") %>% 
  ungroup %>% 
  dplyr::select(bird, `0.5`, fixed_effect) %>% 
  pivot_wider(names_from = fixed_effect, values_from = `0.5`) %>% 
  dplyr::select(-bird) %>% 
  as.matrix() %>% 
  cor %>% 
  corrplot("ellipse", type = "lower", diag = F, addCoef.col = "black", number.digits = 2)

dev.off()
```



We have a look at the parameter estimates for the Alpha parameters (random factor, spatial autocorrelation)

```{r}

lapply(models_post$m2$Alpha[[1]], as.data.frame) %>% 
  bind_rows(.id = "chain") %>% 
  group_by(chain) %>% 
  mutate(sample = 1:n()) %>% 
  pivot_longer(cols=starts_with("A"), names_to = "parameter") %>% 
  ggplot(aes(x = parameter, y = value/1000))+
  geom_violin(draw_quantiles = c(0.5))+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = -90, vjust = 0.5, hjust=1))+
  coord_flip()+
  scale_y_log10()
  
```


```{r}
Gradient = constructGradient(models$m1, focalVariable = "Granskog")
predY = predict(models$m1, Gradient = Gradient, expected = TRUE)

Barrskog <- 
do.call(rbind, predY) %>% 
  as.data.frame() %>% 
  tibble::rowid_to_column() %>% 
  mutate(gradient = rep(Gradient$XDataNew$Granskog, length(predY))) %>% 
  pivot_longer(cols = matches(m1$spNames), names_to = "bird") 

Barrskog_sum <- 
  Barrskog %>% 
  group_by(gradient, bird) %>% 
  summarise(y = quantile(value, 0.5),
            ymin = quantile(value, 0.25),
            ymax = quantile(value, 0.75))
  

BS_plot <- 
ggplot(Barrskog, aes(x = gradient, y = value))+
  geom_point(size = 0.2, alpha = 0.3, position = position_jitter(width = 0.1), colour = "grey")+
  geom_line(data = Barrskog_sum, aes(x = gradient, y = y), colour = "darkred")+
  geom_line(data = Barrskog_sum, aes(x = gradient, y = ymin), colour = "darkblue", size = 0.3)+
  geom_line(data = Barrskog_sum, aes(x = gradient, y = ymax), colour = "darkblue", size = 0.3)+
  facet_wrap(~bird, scales = "free")+
  theme_bw()

BS_plot
```

#predict

import LandUse data for Jönköping (over which we will make predictions)
```{r}
LandUse_JKP <- readRDS( here("data", "JKP_predict_LandUse_data.RDS"))
JKP_points <- readRDS( here("data", "JKP_predict_points.RDS"))
```

recode to match model 

```{r}
LandUse_JKP_X <- 
LandUse_JKP %>% 
  mutate(Lövskog = 
           Ädellövskog +
           Triviallövskog +
           Triviallövskog_med_ädellövinslag,
         Barrskog = 
           Lövblandad_barrskog+
           Tallskog+
           Granskog+
           Barrblandskog,
         Cereal = 
           Spring_cereal+
           Winter_cereal,
           #Tuber_Root,  #not in data
         Open = 
           Fallow+
           Öppen_våtmark+
           Övrig_öppen_mark_med_vegetation+
           Övrig_öppen_mark_utan_vegetation,
         Clearcut = 
           Temporärt_ej_skog
           ) %>% 
  dplyr::select(Lövskog, Barrskog, Clearcut, Pasture, Ley, Cereal, Open, Other)
  
```

spatial data
```{r}
LandUse_JKP_SX <- 
  LandUse_JKP %>% 
  dplyr::select(X,Y)
```

prepare gradient for model fitting
```{r}
# remove first random level (points)
mod2_p <- mod2
mod2$ranLevels$sampling_loc <- NULL
mod2$rL$sampling_loc <- NULL
mod2$rLNames <- "karta" 
mod2$nr <- 1

Gradient = prepareGradient(mod2_test, XDataNew = LandUse_JKP_X, sDataNew = list(karta = LandUse_JKP_SX))

#change class to only data.frame (otherwise function fails)
class(Gradient$XDataNew) <- "data.frame"

#compute predicted values
nParallel=3
predY = predict(mod2_test, Gradient=Gradient, expected = TRUE, nParallel=nParallel)

```


mean parameter estimates
```{r}
EpredY=Reduce("+",predY)/length(predY)
```


```{r}
# expected abundance of the Starling. 
Starling = EpredY[,8]

zero_one <- function(x) {(x - min(x)) / (max(x) - min(x))}

JKP_points %>% 
  mutate(Starling = Starling) %>% 
  ggplot(aes(fill = Starling))+
  geom_sf(colour = "NA")+
  scale_fill_viridis_c()

JKP_points %>% 
  mutate(Starling = Starling) %>% 
  mutate(Pasture = LandUse_JKP_X$Pasture) %>%
  gather(-karta,-geometry, key = "map", value = "value") %>% 
  group_by(map) %>% 
  mutate(value = zero_one(value)) %>% 
  ggplot(aes(fill = value))+
  geom_sf(colour = "NA")+
  scale_fill_viridis_c()+
  facet_wrap(~map)

?gather
```


