---
title: "Hight age"
output: html_notebook
---

I this script we 

+ take the clear cut data from skogsstyrelsen (for which we know the age)
+ take the hight and density layers from the NMD and extract the values for the shapes (at the time of the laserscann)
+ take the dates of the laserscanns to calculate the age of each clearcut at the time of laserscann
+ build a model that predicts the age of clearcut at the time of the laserscann based on hight, density and NA fraction

```{r, message = FALSE}
library(sf)
library(raster)
library(dplyr)
library(tidyr)
library(lubridate)
library(here)
library(ggplot2)
library(future.apply)
```

read in clearcut
read height data (0-5m & 5-45m)
read date of Laserscann data

```{r}
clearcut <- st_read(here("data", "ClearCuts", "sksUtfordAvverk.shp"))

height05 <- raster(here("data", "NMD_object_täckning_o_höjd", "Objekt_hojd_intervall_05_till_5_v1_2", "objekt_hojd_intervall_0_5_till_5_v1_2.img"))

height45 <- raster(here("data", "NMD_object_täckning_o_höjd", "Objekt_hojd_intervall_5_till_45_v1_2", "objekt_hojd_intervall_5_till_45_v1_2.img"))

cover05 <- raster(here("data", "NMD_object_täckning_o_höjd", "Objekt_tackning_hojdintervall_0_5_till_5_v1_2", "objekt_tackning_hojdintervall_0_5_till_5_v1_2.img"))

cover45 <- raster(here("data", "NMD_object_täckning_o_höjd", "Objekt_tackning_hojdintervall_5_till_45_v1_2", "objekt_tackning_hojdintervall_5_till_45_v1_2.img"))

Flygdat <- st_read(dsn = here::here("data", "NMD", "NMD2018metadata_v4.gdb"), 
                   layer = "LM_Laserdata")

#make raster in script
#TNF_vet <- raster(here("data", "NMD", "NMD_CC_polygone", "NMD_tempNofor_vetlanda.tif"))

# only needed later
# CC_raster <- raster(here("data", "ClearCuts", "ClearCuts.tif"))
```



crop raster to common extent and stack it
```{r}

rasterlist <- list(height05, height45, cover05, cover45)

common_extent <- sapply(1:4, function(x) min( unlist( lapply(
  rasterlist, function(y) extent(y)[x])))) %>% 
  extent()
  
croped_rasterlist <- lapply(rasterlist, function(x) crop(x, common_extent))


#stack raster
CC_stack <- stack(croped_rasterlist)

#write stacked raster
writeRaster(CC_stack, filename=here("data", "NMD_object_täckning_o_höjd","stacked_hight_dens.tif"), overwrite=TRUE)


```

for each Kommun, we sample 10% of the clearcuts, extract the data and write them to disk

list of Kommuncodes

```{r}
Kommun_list <- sort(as.character(unique(clearcut$Kommunnr)))

for (i in Kommun_list) {
  
  # check if file allready exist and run code if not
  if( ! file.exists(here("data", "out", "CC_height_data", 
                         paste("ClearCutHeight", i, sep = "_")))) {
    
                         
  # filter for each kommun
  clearcut_filt <- filter(clearcut, Kommun == "Vetlanda") 

  # crop stack to sample extent
  CC_stack_filt <- crop(CC_stack, extent(clearcut_filt))
  
  # make list of clearcut shape
  clearcut_filt_list <-
    clearcut_filt %>%
    split(.$OBJECTID)
  
  ### extarct data for each shape
plan(multiprocess, workers = 16)

CC_height_list <- 
  
 future_lapply(clearcut_filt_list, function(x) {
    
    r <- crop(CC_stack_filt, x) #crop raster to polygon
    
    df <-  raster::extract(r, x) %>%  #extract values and calc summary stat
      as.data.frame() %>% 
      mutate(OBJECTID = x$OBJECTID) %>% 
      rename(height05 = objekt_hojd_intervall_0_5_till_5_v1_2,
             height45 = objekt_hojd_intervall_5_till_45_v1_2,
             cover05 = objekt_tackning_hojdintervall_0_5_till_5_v1_2,
             cover45 = objekt_tackning_hojdintervall_5_till_45_v1_2) %>%
      group_by(OBJECTID) %>%
      summarise( mean_height_05 = mean(height05, na.rm = TRUE),
                 NAs_h_05 = sum(is.na(height05)),
                 mean_height_45 = mean(height45, na.rm = TRUE),
                 NAs_h_45 = sum(is.na(height45)),
                 mean_cover_05 = mean(cover05, na.rm = TRUE),
                 mean_cover_45 = mean(cover45, na.rm = TRUE),
                 size = n())
    })

  #bind rows
  ClearCutHeight <- 
  bind_rows(CC_height_list) 
  
  #write file

  if(! dir.exists(here("data", "out", "CC_height_data"))){
    dir.create(here("data", "out", "CC_height_data"))
    }
  
  write.table(ClearCutHeight,
              here("data", "out", "CC_height_data", 
                   paste("ClearCutHeight", i, sep = "_")),
              sep = "\t")
  
  }
  
}



```


extract flygdatum for each clearcut
```{r}

Flygdat_cc <- clearcut_filt %>% 
  dplyr::select(OBJECTID, Avvdatum, geometry) %>% 
  mutate(geometry = st_centroid(geometry)) %>% 
  st_join(Flygdat)

```



```{r}

clearcut_filt_age <- 
  Flygdat_cc %>% 
  mutate(age = interval(ymd(Avvdatum), ymd(Flygdatum)) %/% days(1)) %>% 
  dplyr::select(OBJECTID, age)

ClearCutHeight <- 
  ClearCutHeight %>% 
  mutate(OBJECTID = as.numeric(OBJECTID)) %>% 
  left_join(clearcut_filt_age) %>% 
  dplyr::select(-geometry)

saveRDS(ClearCutHeight, here::here("data", "ClearCuts", "ClearCutHeightAge.Rds"))

ClearCutHeight <- readRDS(here::here("data", "ClearCuts", "ClearCutHeightAge.Rds"))

mybreaks <-
  c(seq(0, min(ClearCutHeight$age, na.rm = T), -2*365),
    seq(0, max(ClearCutHeight$age, na.rm = T), 2*365)) %>% 
  unique() %>% 
  sort()

mylabels <- 2*seq(sum(mybreaks<0)*-1 , sum(mybreaks > 0), 1)

ClearCutHeight %>%
  mutate(NAs_h_05 = NAs_05/size,
         NAs_h_45 = NAs_45/size) %>% 
  gather(class, value, -OBJECTID, -size, -age) %>% 
  mutate(class = factor(class, levels = unique(class)[c(1,5,2,3,6,4)])) %>% 
  ggplot(aes(x = age, y = value, colour = log(size, base = 10)))+
  geom_point(position = position_jitter(width = 50, height = 0), size = 0.01)+
  facet_wrap(~class, scales = "free_y")+
  geom_smooth(se = FALSE)+
  geom_vline(xintercept = 0)+
  scale_colour_viridis_c()+
  scale_x_continuous(breaks = mybreaks,
                     labels = mylabels)+
  theme_bw()+
  labs(x = "age of clearcut at date of Laserscann (years)")

```



```{r}
# crop clear cut raster to Vetlanda
CC_raster_vet <- crop(CC_raster, TNF_vet)

# recalssify TNF_vet to only one CC class

TNF_vet <- reclassify(TNF_vet, matrix(c(128,1,118,1),ncol = 2, byrow = T))

# stack TNF_vet (which contains only the temporary no forest classes from the NMD)
CC_stack_vet <- stack(TNF_vet, CC_raster_vet)

# function to reclassify TNF_vet to exclude all pixels that overlapp with CC data (i.e. 'have a date')
recl_fun <- function(x,y){ ifelse( is.na(y), x, NA )}

# reclassify
TNF_noage <- overlay(CC_stack_vet, fun = recl_fun)

# export
writeRaster(TNF_noage, here("data", "NMD", "NMD_CC_polygone","TNF_noage.tif"), overwrite=TRUE)

# polygonize with gdal.py
system(
paste(Sys.which("gdal_polygonize.py"),
here("data", "NMD", "NMD_CC_polygone","TNF_noage.tif"),
"-f \"ESRI Shapefile\"",
here("data", "NMD", "NMD_CC_polygone","TNF_noage.shp")))

# read in polygonized TNF
TNF_poly <- st_read(here("data", "NMD", "NMD_CC_polygone","TNF_noage.shp"))

#filter all polygones with less than 10 pixel (1000 m2)
TNF_poly <- 
  TNF_poly %>% 
  mutate(area = st_area(geometry)) %>% 
  filter(as.numeric(area) > 1000) %>% 
  mutate(id = formatC(1:n(), width = nchar(n()), flag = 0))

```

we extract the hight and cover data for the TNF plygones with no age


```{r}
# extract raster from heightraster for each TNF with no age

TNF_list <-
  TNF_poly %>%
  split(.$id)

```


```{r}

# extract rastervalues in for each clearcut for both layers
plan(multiprocess, workers = 4)

options(future.globals.maxSize= 943718400)

TNF_height_list <- 
  
 future_lapply(TNF_list, function(x) {
    
    r <- crop(CC_stack, x) #crop raster to polygon
    
    df <-  raster::extract(r, x) %>%  #extract values and calc summary stat
      as.data.frame() %>% 
      mutate(id = x$id) %>% 
      rename(height05 = objekt_hojd_intervall_0_5_till_5_v1_2,
             height45 = objekt_hojd_intervall_5_till_45_v1_2,
             cover05 = objekt_tackning_hojdintervall_0_5_till_5_v1_2,
             cover45 = objekt_tackning_hojdintervall_5_till_45_v1_2) %>%
      group_by(id) %>%
      summarise( mean_height_05 = mean(height05, na.rm = TRUE),
                 NAs_h_05 = sum(is.na(height05)),
                 mean_height_45 = mean(height45, na.rm = TRUE),
                 NAs_h_45 = sum(is.na(height45)),
                 mean_cover_05 = mean(cover05, na.rm = TRUE),
                 mean_cover_45 = mean(cover45, na.rm = TRUE),
                 size = n())
    })


TNFHeight <- 
  bind_rows(TNF_height_list) 


TNFHeight %>% 
  mutate(NAs_h_05 = NAs_h_05/size,
         NAs_h_45 = NAs_h_45/size) %>% 
  gather(var, val, -id, -size) %>% 
  ggplot(aes(x = val))+
  geom_histogram()+
  facet_wrap(~var, scales = "free")
  
```

```{r}
a <- rbeta(18,2,1)
b <- rbeta(100,1, 89)
#a
#b
sd(a)
sd(b)

var(c(a,b))

sd1 <- sd(a)
sd2 <- sd(b)
m1 <- mean(a)
m2 <- mean(b)
n1 <- length(a)
n2 <- length(b)
  
  
sd_mean <- function(sd1, sd2, n1, n2, m1, m2){
  
  ((n1*(sd1^2+ (m1- sum(c(n1*m1,n2*m2))/(n1+n2))^2))+
    n2*(sd2^2+ (m2- sum(c(n1*m1,n2*m2))/(n1+n2))^2))/
    (sum(n1, n2)-1)
  
  
}

sd_mean(sd(a), sd(b), length(a), length(b), mean(a), mean(b))

(length(a)*(sd(a)^2+ (mean(a)- mean(c(a,b)))^2)+
  length(b)*(sd(b)^2+ (mean(b)- mean(c(a,b)))^2))/
  (sum(length(a), length(b))-1)

```

