---
title: "joint Land Use"
output: html_notebook
---

```{r, message = FALSE}
library(sf)
library(raster)
library(dplyr)
library(tidyr)
library(here)
library(readr)
library(readxl)
library(stringr)
library(ggplot2)
```

## read Data
```{r}
#LandUse <- readRDS(here("data", "LandUse_points_500m.Rds"))
LandUse <- readRDS(here("data", "LandUse_lines_500m.Rds"))
```

### metadata

```{r}
# metadata
NMD <- raster(here("data", "NMD", "nmd2018bas_ogeneraliserad_v1_0.tif")) 


NMD_meta <- 
NMD@data@attributes %>% 
  as.data.frame() %>% 
  dplyr::select(-Opacity, -COUNT) %>% 
  filter(Klass != "") %>% 
  mutate(Klass = iconv(as.character(Klass), "latin1")) %>% 
  rename(NMD = ID)

SJB_meta <- 
  read_xlsx(here("data", "crop_codes_english.xlsx")) %>%
  rename(Landuse_code = `Landuse code`)

#subset for jkp
JKP_karta <- readRDS(here("data", "JKP_karta.RDS"))
```


calculate summary statistics for Agricultural landuse used in line buffers in JKP

(average across all years)
```{r}
SJB_sum <- 
LandUse %>% 
  filter(karta %in% JKP_karta) %>% #subest to JKP for statistics
  dplyr::select(starts_with("SJB"), "n") %>% 
  pivot_longer(-n) %>% 
  filter(!is.na(value)) %>% 
  group_by(value) %>% 
  summarise(n = sum(n)) %>% 
  mutate(prct = signif((n / sum(n)) * 100, 2)) %>% 
  rename(Landuse_code = "value") %>% 
  mutate(Landuse_code = as.numeric(Landuse_code))
```


get unique crop code definition:

SJB_meta contains multiple definition if they changed over the years. However, the most recent definition is from 2012 and all our data are from 2012 onwards. 

some cropcodes are not defined in 2012 but in the data. for these we take the most recent definition from previous years

```{r}
used_codes <- SJB_sum$Landuse_code

#code not defined in 2012
used_codes[!used_codes %in% {filter(SJB_meta, year == "2012") %>% pull(Landuse_code)}]
```

select most recent definition for each cropcode 
```{r}
SJB_meta <- 
SJB_meta %>% 
  arrange(Landuse_code, desc(year)) %>% 
  group_by(Landuse_code) %>% 
  slice(1)

SJB_meta <- 
left_join(SJB_sum, SJB_meta) 

write_tsv(SJB_meta, here("data", "SJB_meta.txt"))
```

check for duplicates and undefined codes
```{r}
SJB_meta %>% 
  group_by(English) %>% 
  filter(n() > 1)
```

```{r}
  SJB_meta %>% 
  arrange(desc(prct)) %>% 
  filter(!is.na(English)) %>% 
  arrange(prct) %>% 
  group_by(English) %>% 
  mutate(prct = sum(prct)) %>% 
  filter(Landuse_code != 62) %>% 
  mutate(English = str_trunc(English, width = 30, "right")) %>%
  mutate(English = factor(English, levels = .$English)) %>% 
  mutate(prct_cum = cumsum(prct)) %>% 
  mutate(col = case_when(prct_cum < 5 & prct_cum >= 1 ~ "less than 5%",
                         prct_cum < 1 ~ "less than 1%",
                         prct_cum > 50 ~ "more than 50%",
                         prct_cum >= 5 & prct_cum <= 50 ~ "more than 95%")) %>% 
  pivot_longer(starts_with("prc")) %>% 
  ggplot(aes(x = English, y = value, colour = col))+
  geom_point()+
  facet_wrap(~name, scales = "free_x")+
  coord_flip()+
  theme_minimal()+
  scale_color_brewer(palette = "Set1")+
  labs(y = "percent (per group and cumulative)",
       x = "crop",
       title = "relative amount of each agriculural land cover - by cropcode")
  
ggsave(filename = here("plots", "crop_code_dist.jpg"), height = 8, width = 8)
```


## recode cropcodes

get crop codes and summarise crop codes to 8 categories

```{r}

# load cropcodes
LandUse_SJB_groups <- SJB_meta


#define groups
LandUse_SJB_groups <-
  LandUse_SJB_groups %>% 
  mutate(crop_groups = case_when(
    Landuse_code %in% c(52,54,55,56,89,95,96) ~ "Pasture",
    Landuse_code %in% c(49,50,51,53,57,58,59,80,81) ~ "Ley",
    Landuse_code %in% c(1,3,4,7,8,12,20,22) ~ "Winter_cereal",
    Landuse_code %in% c(2,5,9,21,23,30,31) ~ "Spring_cereal",
    Landuse_code %in% c(60,62,69) ~ "Fallow",
    Landuse_code %in% c(45, 46,47,32,35) ~ "Tuber_Root", 
    Landuse_code %in% c(65,67,68) ~ "Woody_crops",
    TRUE ~ "Other"))

#recode SJB columns in landuse
LandUse <- 
  LandUse %>% 
  mutate_at(vars(starts_with("SJB_")), function(x) {
    LandUse_SJB_groups[ match( x, LandUse_SJB_groups$Landuse_code),]$crop_groups
  })

#export crop code grouping
LandUse_SJB_groups %>% 
  select(Landuse_code, English, crop_groups, prct) %>% 
  group_by(crop_groups) %>% 
  mutate(group_prct = sum(prct)) %>% 
  arrange(desc(group_prct), desc(prct)) %>% 
  mutate(English = str_trunc(as.character(English), width = 30)) %>% 
  write_tsv(here("crop_groupings.txt"))

#plot grouped landuse
LandUse_SJB_groups %>% 
  select(Landuse_code, English, crop_groups, prct) %>% 
  group_by(crop_groups) %>% 
  mutate(group_prct = sum(prct)) %>% 
  select(crop_groups, group_prct) %>% 
  distinct() %>% 
  arrange(desc(group_prct)) %>% 
  ungroup() %>% 
  mutate(cum_prct = cumsum(group_prct)) %>% 
  arrange(group_prct) %>% 
  mutate(crop_groups = factor(crop_groups, levels = .$crop_groups)) %>% 
  pivot_longer(-crop_groups) %>% 
  ggplot(aes(x = crop_groups, y = value, colour = name))+
  geom_point()+
  geom_bar(data = . %>% filter(name == "group_prct"), stat = "identity", fill = NA)+
  coord_flip()+
  theme_minimal()+
  scale_color_brewer(palette = "Set1")+
  labs(y = "percent (per group and cumulative)",
         x = "crop group",
       title = "relative amount of each agriculural land cover - grouped")


ggsave(filename = here("plots", "crop_groups.jpg"), height = 4, width = 6)  
```

## recode NMD

Distibution of Landuse classes in NMD

```{r}
LandUse %>% 
  filter(karta %in% JKP_karta) %>% 
  group_by(NMD) %>% 
  summarise(n = sum(n)) %>% 
  left_join(NMD_meta) %>% 
  mutate(Klass = ifelse(NMD == 0, "Outside_map", Klass)) %>% 
  mutate(prct = n/sum(n)*100) %>% 
  arrange(desc(prct)) %>% 
  ungroup() %>% 
  mutate(cum_prct = cumsum(prct)) %>% 
  arrange(prct) %>% 
  mutate(Klass = factor(Klass, levels = .$Klass)) %>% 
  select(Klass, contains("prct")) %>% 
  pivot_longer(-Klass) %>% 
  ggplot(aes(x = Klass, y = value, colour = name))+
  geom_point()+
  geom_bar(data = . %>% filter(name == "prct"), stat = "identity", fill = NA)+
  coord_flip()+
  theme_minimal()+
  scale_color_brewer(palette = "Set1")+
  labs(y = "percent (per group and cumulative)",
         x = "crop group")

ggsave(filename = here("plots", "NMD_code_dist.jpg"), height = 6, width = 8)  
```

-> we do not distinguish between forest on wetland and forest on dry land
-> we do not distinguish between 

51  Exploaterad mark, byggnad		
52	Exploaterad mark, ej byggnad eller väg/järnväg		
53	Exploaterad mark, väg/järnväg

recode the NMD column in LandUse

```{r}

LandUse <-
LandUse %>%
  mutate( NMD = case_when(
    NMD %in%  c(121:128) ~ NMD - 10,
    NMD %in%  c(52,53) ~ 51,
    TRUE ~ NMD))

```


Distribution of grouped Landuse classes in NMD
```{r}
LandUse %>% 
  filter(karta %in% JKP_karta) %>% 
  group_by(NMD) %>% 
  summarise(n = sum(n)) %>% 
  left_join(NMD_meta) %>% 
  mutate(Klass = ifelse(NMD == 0, "Outside_map", Klass)) %>% 
  mutate(Klass = gsub("(.+)\\(utanför våtmark\\)", "\\1", Klass)) %>% 
  mutate(prct = n/sum(n)*100) %>% 
  arrange(desc(prct)) %>% 
  ungroup() %>% 
  mutate(cum_prct = cumsum(prct)) %>% 
  arrange(prct) %>% 
  mutate(Klass = factor(Klass, levels = .$Klass)) %>% 
  select(Klass, contains("prct")) %>% 
  pivot_longer(-Klass) %>% 
  ggplot(aes(x = Klass, y = value, colour = name))+
  geom_point()+
  geom_bar(data = . %>% filter(name == "prct"), stat = "identity", fill = NA)+
  coord_flip()+
  theme_minimal()+
  scale_color_brewer(palette = "Set1")+
  labs(y = "percent (per group and cumulative)",
         x = "crop group")

ggsave(filename = here("plots", "NMD_groups.jpg"), height = 4, width = 6)  
```

## join Landuse

The NMD contains the LandUse class `3 (Åkermark)` onto which we match the data from SJB. 

However the match is not perfect (because the NMD has taken the blockdata - not the application data as template and because & because the exact polygones change a bit between the years while the NMD does not)

We let the SJB data take over the NMD data 

```{r}
LandUse %>% 
  filter(karta %in% JKP_karta) %>% 
  pivot_longer(starts_with("SJB"), names_to = "year", values_to = "Crop") %>% 
  group_by( year, Crop, NMD) %>% 
  summarise(n = sum(n)) %>% 
  group_by(year, Crop) %>% 
  mutate(n = round(n / sum(n) * 100, 1)) %>% 
  left_join(NMD_meta) %>% 
  filter(Klass == "Åkermark") %>% 
  pivot_wider(names_from = year, values_from = n) %>% 
  select(-NMD)
  
```
The one class that doesn't match the class `3 (Åkermark)` is `Pasture`. This is because `Pasture` is an additional layer in the NMD. 

```{r}
LandUse %>% 
  filter(karta %in% JKP_karta) %>% 
  filter(SJB_2012 == "Pasture") %>% 
  group_by(SJB_2012, NMD) %>% 
  summarise(n = sum(n)) %>% 
  arrange(desc(n)) %>% 
  mutate(n = round(n / sum(n) * 100, 1)) %>% 
  left_join(NMD_meta)
```

Pasture is split between pasture with some sort of tree cover and pasture without tree cover

```{r}

LandUse %>% 
  filter(karta %in% JKP_karta) %>% 
  left_join(NMD_meta) %>% 
  filter(SJB_2012 == "Pasture") %>% 
   left_join(NMD_meta) %>% 
  mutate(SJB_2012 = case_when(SJB_2012 == "Pasture" & grepl("\\wskog", Klass) ~ paste(SJB_2012, "with_trees", sep = "_"),
                              TRUE ~ SJB_2012)) %>% 
  group_by(SJB_2012) %>% 
  summarise(n = sum(n)) %>% 
  arrange(desc(n)) %>% 
  mutate(`%` = round(n / sum(n) * 100, 1))

```

1) We let the SJB take over the NMD
2) We split Pasture in 2 classes (with and without forest, according to NMD)
3) SJB Landuse class `Woody_crops` is replaced by class `115	Triviallövskog (utanför våtmark)`

Not all pixel that are coded `3 Åkermark`in the NMD are covered by the SJB. It's seldom much (see above) but we need to deal with it. 

4) We declare the remaining `Åkermark` as `Övrig_öppen_mark_med_vegetation`


5) We add a class for NMD = 0 which codes for pixels outside the border of the map (during extraction from buffers - not relevant for JKP)

create new NMD metadata with grouped landuse classes
```{r}


LandUse_meta <- 
NMD_meta %>% 
  filter(NMD %in% unique(LandUse$NMD)) %>% 
  rename(LandUse = Klass,
         LandUse_code = NMD) %>% 
  #collapse forest classes
  mutate(LandUse = gsub("(\\w+)\\s\\(utanför våtmark\\)", "\\1", LandUse), 
         #collapse building classes
         LandUse = gsub("(\\w+), byggnad", "\\1", LandUse),
         #strip remaining spaces
         LandUse = gsub("\\s", "_", LandUse)) %>% 
  #add class for code 0
  rbind(tibble(LandUse_code = 0, LandUse = "Outside_map" ))
```

reclassify NMD

1) change NMD codes to factor level
2) recode NMD data with SJB
```{r}
LandUse_long <- 
LandUse %>% 
  mutate(rowID = 1:n()) %>% 
  mutate(NMD = LandUse_meta[ match(NMD, LandUse_meta$LandUse_code),]$LandUse) %>% 
  pivot_longer(starts_with("SJB")) %>% 
  mutate(NMD = case_when(value == "Pasture" & grepl("\\wskog", NMD) ~ "Pasture_w_Trees",
                         value == "Woody_crops" ~ "Triviallövskog",
                         !is.na(value) ~ value,
                         TRUE ~ NMD)) 


```


Not all pixel that are coded `3 Åkermark`in the NMD are covered by the SJB. It's seldom much but we need to deal with it. 

we declare the remaining `Åkermark` as `Övrig_öppen_mark_med_vegetation`. 

```{r}
LandUse_long <- 
LandUse_long %>% 
  mutate(NMD = case_when(NMD == "Åkermark" ~ "Övrig_öppen_mark_med_vegetation",
                         TRUE ~ NMD))
```

LandUse code 0 ("Outside_map") is for pixel outside NMD. It needs to be dealt with.

```{r}
Routes_not_covered <- 
LandUse_long %>% 
  select(karta, position, NMD, n) %>% 
  filter(NMD == "Outside_map") %>% 
  distinct() %>% 
  group_by(karta, position) %>% 
  mutate(n = sum(n)) %>% 
  group_by(karta) %>% 
  mutate(N_Excl = n())

```

I exclude all routes which have >= 4 points that are not fully covered by the NMD. 
For the remaining routes I exclude only the buffers with missing data.

```{r}

LandUse_long <-  
LandUse_long %>%  
  filter( !karta %in% filter(Routes_not_covered, N_Excl >= 4)$karta) %>% 
  filter( !paste(karta, position) %in% 
            paste(filter(Routes_not_covered, N_Excl <4)$karta,
                  filter(Routes_not_covered, N_Excl <4)$position)) 

```

Distribution of joined Landuse classes
```{r}
LandUse_long %>% 
  filter(karta %in% JKP_karta) %>% 
  group_by(NMD, name) %>% 
  summarise(n = sum(n)) %>% 
  group_by(name) %>% 
  mutate(prct = n/sum(n)*100) %>% 
  arrange(desc(name), prct) %>% 
  ungroup() %>% 
  mutate(NMD = factor(NMD, levels = unique(.$NMD))) %>%
  mutate(name = factor(name, levels = unique(.$name))) %>% 
  ggplot(aes(x = NMD, y = prct, fill = name))+
  geom_bar(position = position_dodge(width = 0.8), stat = "identity")+
  coord_flip()+
  theme_minimal()+
  scale_fill_brewer(palette = "Paired")#+
  labs(y = "percent (per group and cumulative)",
         x = "crop group")

ggsave(filename = here("plots", "LandUse_groups.jpg"), height = 6, width = 8)  
```

## export
transform back to wide format

```{r}
LandUse <- 
  LandUse_long %>% 
  select(-value) %>% 
  group_by(karta, position, name, NMD) %>% 
  summarise(n = sum(n)) %>% 
  pivot_wider(names_from = NMD, values_from = n, values_fill = 0) %>% 
  rename(year = name)
  
```


```{r}
#write_csv(LandUse, here("data", "LandUse_joint_points.csv"))
write_csv(LandUse, here("data", "LandUse_joint_lines.csv"))
```


#QGIS

make colour maps for QGIS project

##NMD

```{r}
NMD_cols <- read_delim(here("NMD_codes.clr"), delim = " ", 
                       col_names = c("Landuse_code", "X1", "X2", "X3", "X4", "Label"))
NMD_cols <- 
NMD_cols %>% 
  left_join(dplyr::select(SJB_meta, Landuse_code, English)) %>% 
   mutate(crop_groups = case_when(
    Landuse_code %in% c(52,54,55,56,89,95,96) ~ "Pasture",
    Landuse_code %in% c(49,50,51,53,57,58,59,80,81) ~ "Ley",
    Landuse_code %in% c(1,3,4,7,8,12,20,22) ~ "Winter_cereal",
    Landuse_code %in% c(2,5,9,21,23,30,31) ~ "Spring_cereal",
    Landuse_code %in% c(60,62,69) ~ "Fallow",
    Landuse_code %in% c(45, 46,47,32,35) ~ "Tuber_Root", 
    Landuse_code %in% c(65,67,68) ~ "Woody_crops",
    is.na(English) ~ NA_character_,
    TRUE ~ "Other"))

cols <- 
NMD_cols %>% 
  group_by(crop_groups) %>% 
  slice(1) %>% 
  dplyr::select(crop_groups, starts_with("X"))

NMD_cols <- 
NMD_cols %>% 
 left_join(cols, by = "crop_groups") %>% 
 dplyr::select(Landuse_code, ends_with("y"), crop_groups)


write_delim(NMD_cols, file = here("NMD_codes_new.clr"), col_names = FALSE)
```

