---
title: "joint Land Use"
output: html_notebook
---

```{r, message = FALSE}
library(sf)
library(raster)
library(dplyr)
library(tidyr)
library(here)
library(readr)
library(readxl)
library(stringr)
library(vegan)
```

### read in LandUse type from different source Raster
```{r}
LandUse <- readRDS(here("data", "LandUse_points_500m.Rds"))
#LandUse <- readRDS(here("data", "LandUse_lines_500m.Rds"))
```

### read in metadata

```{r}
# metadata
NMD <- raster(here("data", "NMD", "nmd2018bas_ogeneraliserad_v1_0.tif")) 


NMD_meta <- 
NMD@data@attributes %>% 
  as.data.frame() %>% 
  dplyr::select(-Opacity, -COUNT) %>% 
  filter(Klass != "") %>% 
  mutate(Klass = iconv(as.character(Klass), "latin1")) %>% 
  rename(NMD = ID)

#this selects the most recent definition of the crop_code if they have been redefined over the years. However, the meaning of the crop code doesn't change substantially in most cases and in no major case

SJB_meta <- 
  read_xlsx(here("data", "crop_codes_english.xlsx")) %>% 
  rename(Landuse_code = `Landuse code`) %>% 
  dplyr::select(Landuse_code, English) %>% 
  distinct() %>% 
  group_by(Landuse_code) %>% 
  filter(row_number()==n())

```


#### recode

get cropcodes and summarise cropcodes to 8 categories

recode cropcodes in LandUse data

```{r}

# load cropcodes
LandUse_SJB_groups <- 
  read_csv2(here("data", "CropCodes.csv"))

#define groups
LandUse_SJB_groups <-
  LandUse_SJB_groups %>% 
  mutate(crop_groups = case_when(
    Landuse_code %in% c(52,54,55,56,89,95,96) ~ "Pasture",
    Landuse_code %in% c(49,50,51,53,57,58,59,80,81) ~ "Ley",
    Landuse_code %in% c(1,3,4,7,8,12,20,22) ~ "Winter_cereal",
    Landuse_code %in% c(2,5,9,21,23,30,31) ~ "Spring_cereal",
    Landuse_code %in% c(60,62,69) ~ "Fallow",
    Landuse_code %in% c(45, 46,47,32,35) ~ "Tuber_Root", 
    Landuse_code %in% c(65,67,68) ~ "Woody_crops",
    TRUE ~ "Other"))

#recode SJB columns in landuse
LandUse <- 
  LandUse %>% 
  mutate_at(vars(starts_with("SJB_")), function(x) {
    LandUse_SJB_groups[ match( x, LandUse_SJB_groups$Landuse_code),]$crop_groups
  })

```

## recode Landuse classes from NMD
-> we do not distinguish between forest on wetland and forest on dry land
-> we do not distinguish between 

51  Exploaterad mark, byggnad		
52	Exploaterad mark, ej byggnad eller väg/järnväg		
53	Exploaterad mark, väg/järnväg

recode the NMD column in LandUse

```{r}

LandUse <-
LandUse %>%
  mutate( NMD = case_when(
    NMD %in%  c(121:128) ~ NMD - 10,
    NMD %in%  c(52,53) ~ 51,
    TRUE ~ NMD))

```


+ NMDbase landuse class `3 (Åkermark)` is filled in by data from SJB
+ SJB Landuse class `Woody_crops` is replaced by class `115	Triviallövskog (utanför våtmark)`

+ we add a class for NMD = 0 which codes for pixels outside the border of the map (during extraction from buffers)

```{r}

LandUse_meta <- 
NMD_meta %>% 
  filter(NMD %in% unique(LandUse$NMD)) %>% 
  rename(LandUse = Klass,
         LandUse_code = NMD) %>% 
  mutate(LandUse = gsub("(\\w+)\\s\\(utanför våtmark\\)", "\\1", LandUse),
         LandUse = gsub("(\\w+), byggnad", "\\1", LandUse),
         LandUse = gsub("\\s", "_", LandUse)) %>% 
  rbind(tibble(LandUse_code = 0, LandUse = "Outside_map" ))

LandUse <- 
LandUse %>% 
  mutate(NMD = LandUse_meta[ match(NMD, LandUse_meta$LandUse_code),]$LandUse) %>% 
  mutate_at(vars(starts_with("SJB_")), list(
    ~case_when(NMD != "Åkermark" ~ NMD,
               . == "Woody_crops" ~ "Triviallövskog",
                  TRUE ~ .))
  )

```


Not all pixel that are coded `3 Åkermark`in the NMD are covered by the SJB. It's seldom much but we need to deal with it. 

we declare the remaining `Åkermark` as `Övrig_öppen_mark_med_vegetation`. 

```{r}
LandUse <- 
LandUse %>% 
  mutate_at(vars(starts_with("SJB_")), list(
    ~case_when(NMD == "Åkermark" & is.na(.) ~ "Övrig_öppen_mark_med_vegetation",
                  TRUE ~ .))
  )
```

LandUse code 0 ("Outside_map") is for pixel outside NMD. It needs to be dealt with.

```{r}

Routes_not_covered <- 
LandUse %>% 
  select(karta, position, NMD, n) %>% 
  group_by(karta, position) %>% 
  filter(NMD == "Outside_map") %>% 
  mutate(n = sum(n)) %>% 
  group_by(karta) %>% 
  mutate(N_Excl = n())

```

I exclude all routes which have >= 4 points that are not fully covered by the NMD. 
For the remaining routes I exclude only the buffers with missing data.

```{r}

LandUse <-  
LandUse %>%  
  filter( !karta %in% filter(Routes_not_covered, N_Excl >= 4)$karta) %>% 
  filter( !paste(karta, position) %in% 
            paste(filter(Routes_not_covered, N_Excl <4)$karta,
                  filter(Routes_not_covered, N_Excl <4)$position)) 
```

## wide format

```{r}
LandUse <- 
LandUse %>% 
  select(-NMD) %>% 
  pivot_longer(cols = starts_with("SJB"), names_to = c("year"), values_to = "LandUse") %>% 
  group_by(karta, position, year, LandUse) %>% 
  summarise(n = sum(n))

LandUse <- 
LandUse %>% 
  pivot_wider(names_from = LandUse, values_from = n, values_fill = 0)

LandUse
```

```{r}
write_csv(LandUse, here("data", "LandUse_joint_points.csv"))
#write_csv(LandUse, here("data", "LandUse_joint_lines.csv"))
```


