---
title: "Jönköping karta"
output: html_notebook
---

```{r}
library(ggplot2)
library(sf)
library(here)
library(dplyr)
library(raster)
library(plotly)
```

```{r}
Point_counts <- 
  read_sf(here("Standardrutter-shapefiler", "stdruttPunkter.shp")) %>% 
  st_set_crs(3021) %>% 
  st_transform(crs = 3006) %>% 
  group_by(KARTA) %>% 
  summarise() %>% 
  st_centroid()
```


# Jönköping region
Here we subset the data to include the Jönköping sample region routes in a similar landscape. 

## define region

Here we define the region for the model 
+ Taking the JKP study region
+ Taking the agricultural production regions defined by SCB
+ Choosing all routes that fall into production region 5 (Götalands Skogsbygder)
+ excluding coastal routes, defined as being within 5 km from the coast


The production areas are :

1 : Götalands Södra Slättbygder
2 : Götalands Mellanbygder
3 : Götalands Norra Slättbygder
4 : Svealands Slätbygder
5 : Götalands Skogsbygder
6 : Mellersta Svergies Skogsbygder
7 : Nedre Norrland
8 : Övre Norrland

```{r}
#production areas
Prod_reg <- 
  read_sf(here("data", "prodzon2.shp")) %>% 
  mutate(Area = case_when(
    PO8 == 1 ~ "1 - Götalands Södra Slättbygder", 
    PO8 == 2 ~ "2 - Götalands Mellanbygder",
    PO8 == 3 ~ "3 - Götalands Norra Slättbygder",
    PO8 == 4 ~ "4 - Svealands Slätbygder",
    PO8 == 5 ~ "5 - Götalands Skogsbygder",
    PO8 == 6 ~ "6 - Mellersta Svergies Skogsbygder",
    PO8 == 7 ~ "7 - Nedre Norrland",
    PO8 == 8 ~ "8 - Övre Norrland"
  )) %>% st_set_crs(3021) %>% 
  st_transform(crs = 3006)

ggplot(Prod_reg)+
  geom_sf(alpha = 0.5, aes(fill = Area))+
  scale_fill_brewer(palette = "Paired")+
  theme_bw()
```

add our study region and the buffer of -5km around the coastline
```{r}
#study region as defined for F2F
JKP <- 
read_sf(here("data", "prod_län_combi_byyann_skarning2.shp")) %>% 
  st_set_crs(3006)


#get country boardes for JKP
SWE <- getData('GADM', country='SWE', level=0)

SWE <- 
SWE %>% 
  st_as_sf() %>%
  st_transform(crs = 3006) %>% 
  st_simplify() %>% 
  st_buffer(-5000)
```

subset the routes that fall within the production region 5 and with > 5 km distance from the coast
```{r}
Point_counts_sel <- 
  Point_counts %>% 
  st_join(filter(Prod_reg, PO8 == 5), join = st_covered_by, left = FALSE) %>% 
  st_join(SWE, join = st_covered_by, left = FALSE)

p <- ggplot(Prod_reg)+
  geom_sf(alpha = 0.5, aes(fill = Area))+
  scale_fill_brewer(palette = "Paired")+
  geom_sf(data = JKP, fill = "white", alpha = 0.5, colour = "red")+
  geom_sf(data = Point_counts, shape = 22)+
  geom_sf(data = Point_counts_sel, colour = "red", shape = 22)+
  geom_sf(data = SWE, colour = "darkblue", fill = NA)+
  theme_bw()
  
p %>% ggplotly

p +
  scale_y_continuous(limits = c(6100000,6600000))+
  scale_x_continuous(limits = c(260000,740000))

ggsave(here("plots", "route_selection.png"))
```

export routes

```{r}
#export list of routes in JKP region
JKP_karta <- Point_counts_sel$KARTA
saveRDS(JKP_karta, here("data", "JKP_karta.RDS"))
```