---
title: "Rasterize the application data (2012-2018)"
output: html_notebook
---

R code to check for abandonment of agricultural blocks and their transformation, especially afforestation. 

```{r, message = FALSE}
library(sf)
library(raster)
library(fasterize)
library(tidyr)
library(dplyr)
library(here)
library(future.apply)
```

load NMD
```{r}
NMD <- raster(here("data", "NMD", "nmd2018bas_ogeneraliserad_v1_0.tif"))
```

function to subdivide the large NMD raster into smaller raster

`ext` takes the raster extent of the original raster
`n_x` takes the number of tiles in x direction that the raster should be divided into
`n_y` takes the number of tiles in y direction that the raster should be divided into

the function gives a list of length n_x * n_y with extent objects covering the original raster. The list can be used to crop the raster into smaller raster


(I also found the `SpaDES.tools::splitRaster` function which does the same than this handcrafted approach)

```{r}

divide_raster <- function(ext, n_x, n_y){
  
  x_vals <- seq(ext@xmin, ext@xmax, length.out = n_x+1)
  y_vals <- seq(ext@ymin, ext@ymax, length.out = n_y+1)
  
  x_vals_list <- x_vals[c(1, rep(2:n_x, each = 2), n_x+1)] %>% split(rep(1:n_x, each = 2))
  y_vals_list <- y_vals[c(1, rep(2:n_y, each = 2), n_y+1)] %>% split(rep(1:n_y, each = 2))
  
  crossing(x_vals_list, y_vals_list) %>% 
  apply(1, function(x) extent(x$x_vals_list, x$y_vals_list))
  
}

```


crop the NMD raster into 25 tiles

```{r}
dir.create(here("data", "NMD", "NMD_tiles"))

plan(multiprocess, workers = 4)

ext_list <- divide_raster(extent(NMD), 10, 10)

future_sapply(
  seq_along(ext_list), function(x) {
    crop(NMD, ext_list[[x]], 
         filename = here("data", "NMD", "NMD_tiles", 
                         paste("NMD_tile_", formatC(x, width = 2, flag = 0), 
                               ".tif", sep = "")))
  }
  )

```

we need to split the shape file into chunks that fit the extent of the raster tiles 
as a shape can fall on the border between two or more raster tiles, we 
```{r}

fasterize_tiles <- function(Tiles, SF, path, ncores){
  
  #Tiles = list of raster
  #SF = simple feature polygones df to rasterize
  #path = were to write the rasterized SF to
  
  plan(multiprocess, workers = ncores)
  
  future_sapply(seq_along(Tiles), function(x){
    
    ext <- extent(Tiles[[x]])
    shp <- filter(SF, xmax >= ext@xmin & xmin <= ext@xmax & ymax >= ext@ymin & ymin <= ext@ymax)
    
    field <- c("GRDKOD_MAR", "MYGRODKOD") #the field changed name in 2015
   
    
     if(nrow(shp) > 0) { 
    shp_r <- fasterize(sf = shp, 
                       raster = Tiles[[x]], 
                       field = field[ which( field %in% colnames( SF))], 
                       fun = "first", 
                       background = NA_real_) 
    
    writeRaster(shp_r, filename = paste(path, paste(deparse(substitute(SF)), x, ".tif", sep = ""), sep ="/"))
    rm(shp_r)}
    return(NULL)
  }
  
  )
  
}
        

```


fasterize first parcel data (2012)
```{r}
#list of raster tiles
NMD_list <- list.files(here("data", "NMD", "NMD_tiles"))
NMD_list <- NMD_list[grepl(".tif$",NMD_list)]
NMD_list <- lapply(NMD_list, function(x) raster(here("data", "NMD", "NMD_tiles", x)))

#path to shapefiles shapefiles
path <- "/Volumes/Fabian_Work/Sverige/Application_data/"
files <- list.files(path)

#path for rasterized shapefiles
path.out <- here("data", "NMD", "Rastershapes")

dir.create(path.out)

#load first shapefile
block2012 <- st_read(paste0(path, files[[1]]))

#fasterize in tiles in parallel

fasterize_tiles(Tiles = NMD_list, SF = block2012, path = path.out, ncores = 4)

```

check if all files have colums for the bounding blocks of each polygon, if not add and re-save

```{r}

#path to shapefiles shapefiles
path <- "/Volumes/FABIAN_USB/Application_data/"

plan(multiprocess, workers = 4)
  

files <- list.files(path)

for (i in files) {
  
 SF <-  st_read(paste0(path, i))
 
  if(!"xmax" %in% colnames(SF)){
    
    g <- st_geometry(SF)
    
    g_bbox <- 
      future_lapply(g, st_bbox) %>% 
      do.call("rbind", .) %>% 
      as_tibble() 
    
    SF_new <- cbind(as_tibble(SF), g_bbox) %>% 
      st_as_sf()
  
    write_sf(SF_new, paste0(path, i))
    
  }
  
}

```



fasterize remaining parcel data (2013-2018)
```{r}
#list of raster tiles
NMD_list <- list.files(here("data", "NMD", "NMD_tiles"))
NMD_list <- NMD_list[grepl(".tif$",NMD_list)]
NMD_list <- lapply(NMD_list, function(x) raster(here("data", "NMD", "NMD_tiles", x)))

#path to shapefiles shapefiles
path <- "/Volumes/Fabian_Work/Sverige/Application_data/"
files <- list.files(path)

for(i in files){
  
  SF <- st_read(paste0(path, i))
  
  year <- gsub("(\\d+).gpkg", "\\1", i)
  
  path.out <- here("data", "NMD", "Rastershapes", year)
  
  dir.create(path.out)
  
  fasterize_tiles(Tiles = NMD_list, 
                  SF = SF, 
                  path = path.out, 
                  ncores = 4)
  
}

```


## merge raster

```{r}

rasterS <- list.files(here("data","NMD","Rastershapes"))

plan(multiprocess, workers = 4)

future_lapply(rasterS, function(x) {
  path_merge <- here("data","NMD","Rastershapes", x)
  files <- list.files(path_merge, pattern = ".tif$")
  merge.list <- lapply(paste(path_merge,files,sep = "/"), raster)
  do.call(merge, c(merge.list, filename = paste0(here("data","SJB", x), ".tif")))
})

plan(sequential)
```

## recode raster

because I didn't pay attention to the fact that fasterize will take the *factor level* and not the numerical value of the LandUse code (because I didn't convert it to a numeric variable first), we need to recode the raster values. 

for that we read in the application shapefiles, create a conversion matrix from factor level to numerical value and recode the raster accordingly

```{r}

path_to_shapes <- "/Volumes/Fabian_Work/Sverige/Application_data"
shapefiles <- list.files(path_to_shapes, pattern = ".gpkg$")

field <- c("GRDKOD_MAR", "MYGRODKOD")

plan(multiprocess, workers = 4)

conv_list <- 
  future_lapply(shapefiles, function(x) {
  
  SHAPE <- st_read(paste(path_to_shapes, x, sep = "/"))
  
  field <- field[ which( field %in% colnames(SHAPE))]
  
  tibble( from = as.numeric(as.factor(unique(as.data.frame(SHAPE)[,field]))),
          to = as.numeric(as.character(unique(as.data.frame(SHAPE)[,field]))))
  
    })

raster_files <- list.files(here("data", "SJB"))
raster_list <- lapply(raster_files, function(x) raster(here("data", "SJB", x)))

plan(multiprocess, workers = 4)

future_lapply(seq_along(raster_list), function(x){
  
reclassify( x = raster_list[[x]], 
            rcl = as.matrix(conv_list[[x]]),
            filename = here("data", "SJB", paste("recl", raster_files[x], sep = "")))  
  
})


```


