---
title: "SJB_characterization"
output: html_notebook
---
## load libraries

```{r, message = FALSE}
library(sf)
library(raster)
library(dplyr)
library(tidyr)
library(here)
library(future.apply)
library(pbapply)
library(readxl)
library(ggthemes)
```

## load bird routes

```{r}
Standard_geo_p <- read_sf(here("data", "Standardrutter-shapefiler", "stdruttPunkter.shp"))
Standard_geo_l <- read_sf(here("data", "Standardrutter-shapefiler", "stdruttLinjer.shp"))
#Standard_geo <- read_sf(here("data", "Standard_routes_geo.gpkg"))
#Freecounts_geo <- read_sf(here("data", "Standard_routes_geo.gpkg"))

Standard_geo_p <- 
  Standard_geo_p %>% 
  dplyr::select(KARTA, PUNK) %>% 
  rename(karta = KARTA, position = PUNK) %>% 
  mutate(type = "point")

Standard_geo_l <- 
  Standard_geo_l %>% 
  dplyr::select(KARTA, LINJE) %>% 
  rename(karta = KARTA, position = LINJE) %>% 
  mutate(type = "line")

Standard_geo <- 
  bind_rows(Standard_geo_p, Standard_geo_l) %>% 
  mutate(position = tolower(position))
```

To keep the information about what LandUse is in what raster in which year, we need to extract from all raster simultaneously. The SJB raster have been rasterized using the NMD as template so all SJB raster as well as the NMD have the same extent and can be stacked. 


## load application data

```{r}

SJB_files <- list.files(here("data", "SJB"), pattern = ".+tif$")
SJB_raster <- pblapply(SJB_files, 
                    function(x) raster(paste(here("data", "SJB"), x, sep = "/")))


gsub("(\\w+).tif", "\\1", SJB_files) %>% 
  paste("SJB", . , sep = "_") -> names(SJB_raster)

```

## load NMD data

```{r}
NMD <- raster(here("data", "NMD", "nmd2018bas_ogeneraliserad_v1_0.tif"))

NMD_meta <- 
  NMD@data@attributes %>% 
  as.data.frame() %>% 
  mutate(Klass = iconv(as.character(Klass), "latin1"))

levels(NMD) <- NMD_meta

NMD_meta <-
  NMD_meta %>% 
  dplyr::select(-Opacity) %>% 
  filter(Klass != "") 
```

stack raster

```{r}

SJB_raster$NMD <- NMD

LandUse_stack <- stack(SJB_raster, NMD)

```

reproject standard geo

```{r}
st_crs(Standard_geo) <-  "EPSG:2400"
Standard_geo <- st_transform(Standard_geo, st_crs(NMD))
```


**make buffer for raster extraction** 
```{r}

# make a 1km buffer around each route
Standard_geo_buffer <- 
  Standard_geo %>% 
  group_by(karta) %>% 
  summarise(.groups = "drop") %>% 
  st_convex_hull() %>% 
  st_buffer(1000) 

```


**crop list of raster for LandUse data**

we make a list of length 7 for the 7 years of SJB application data we have. Each entry is itself a list 
```{r}

# extract raster from SJB for each route for each year
Buffer_list <- 
  Standard_geo_buffer %>% 
  split(.$karta) 

LandUse_stack_list <- 
pblapply(Buffer_list, function(x) {
  crop(LandUse_stack, x)
  })

```


extract raster values in 500 m buffer around points for all routes
```{r}
# vector of route names 
Routes <- names(Buffer_list)

LandUse_point_buffer <- 
  
  pblapply(Routes, function(x) {
    
    filter(Standard_geo, karta == x, type == "point") %>%
      raster::extract(x = LandUse_stack_list[[x]], y =  ., buffer = 500, 
                      cellnumbers = F, df = T) %>%
      as_tibble() %>%
      mutate(karta = x, position = paste("p", ID, sep = "")) %>% 
      group_by(across(c(position, karta, NMD, starts_with("SJB"))))%>% 
      summarise(n = n(), .groups = "drop")
    }) %>% 
  bind_rows()

```

export data
```{r}
saveRDS(LandUse_point_buffer, here("data", "LandUse_points_500m.Rds"))
```

#line counts

## Make buffers 
```{r}

#distance centroid to mid-line point
d <- 1000

# buffer width
b <- 500

Line_buffers <- 
  Standard_geo %>% 
  group_by(karta) %>% 
  summarise(.groups = "drop") %>% 
  st_centroid()%>% 
    mutate(p1 = geometry + c(0,d-b/2),
           p2 = geometry + c(0,d+b/2),
           p3 = geometry + c(d-b/2,d-b/2),
           p4 = geometry + c(d+b/2, d+b/2),
           p5 = geometry + c(d-b/2,0),
           p6 = geometry + c(d+b/2,0),
           p7 = geometry + c(d-b/2,-(d-b/2)),
           p8 = geometry + c(d+b/2,-(d+b/2)),
           p9 = geometry + c(0,-(d-b/2)),
           p10 = geometry + c(0,-(d+b/2)),
           p11 = geometry + c(-(d-b/2),-(d-b/2)),
           p12 = geometry + c(-(d+b/2), -(d+b/2)),
           p13 = geometry + c(-(d-b/2),0),
           p14 = geometry + c(-(d+b/2),0),
           p15 = geometry + c(-(d-b/2),d-b/2),
           p16 = geometry + c(-(d+b/2), d+b/2)) %>% 
    pivot_longer(cols = starts_with("p"), names_to = "point") %>% 
    dplyr::select(-geometry) %>% 
    rename(geometry = value) %>% 
    group_by(karta) %>% 
    slice(c(1,2,3,4,3,4,5,6,5,6,7,8,7,8,9,10,9,10,11,12,11,12,13,14,13,14,15,16,15,16,1,2)) %>%
    mutate(line = rep(rep(c(4:8,1:3), each = 4),n()/32)) %>% 
    st_as_sf()  %>% 
    group_by(karta, line) %>% 
    summarise(.groups = "drop") %>% 
    st_convex_hull() %>% 
    st_set_crs(crs(Standard_geo)) %>% 
    mutate(line = paste("l", line, sep = ""))


Line_buffers %>% 
  filter(karta == "02C2H") %>% 
  ggplot()+
  geom_sf(aes(fill = line))+
  geom_sf(data = Standard_geo %>% filter(karta == "02C2H") %>%  filter(type == "line"))
```

example of line buffer

```{r}

NMD_meta <- 
  NMD_meta %>% 
  mutate(Colour = c(
  "#30b092","#9e6b05","#e3e854","#afc244","#911f1f","#757373","#000000",
  "#287ebf","#063780","#044a04","#036303","#0b5e0b","#147314","#49b849",
  "#009c00","#239623","#819623","#044a3b","#03634b","#0b5e4f","#147355",
  "#49b88c","#009c63","#239677","#7d9623"
  ))

pal = c(NMD_meta$Colour)
names(pal) <- NMD_meta$Klass

NMD_crop <- 
Standard_geo_buffer %>% 
  filter(karta == "02C2H") %>% 
  crop(NMD, .) 

NMD_crop %>% 
  as(., "SpatialPixelsDataFrame") %>% 
  as.data.frame() %>% 
  rename(ID = nmd2018bas_ogeneraliserad_v1_0) %>%
  left_join(NMD_meta) %>% 
  ggplot() +  
  geom_raster(aes(x=x, y=y, fill=Klass, alpha = 1))+
  geom_sf(data = filter(Standard_geo, karta == "02C2H"), colour = "black")+
  geom_sf(data = filter(Line_buffers, karta == "02C2H"), 
          fill = NA, colour = "red")+
  scale_fill_manual(values = pal)+
  theme_map() +
  theme(legend.position = "none")+
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0))


ggsave(here("plots", "linebuffer.jpeg"))
  
```


extract raster values in 250 m buffer around points for all routes
```{r}
# vector of route names 
Routes <- names(Buffer_list)

LandUse_line_buffer <- 
  
  pblapply(Routes, function(x) {
    
    filter(Line_buffers, karta == x) %>%
      raster::extract(x = LandUse_stack_list[[x]], y =  ., cellnumbers = F, df = T) %>%
      as_tibble() %>%
      mutate(karta = x, position = paste("l", ID, sep = "")) %>% 
      group_by(across(c(position, karta, NMD, starts_with("SJB"))))%>% 
      summarise(n = n(), .groups = "drop")
    }) %>% 
  bind_rows()

```


export data
```{r}
saveRDS(LandUse_line_buffer, here("data", "LandUse_lines_500m.Rds"))
```