---
title: "Standard routes descriptive"
output: html_notebook
---

load libraries
```{r, message = FALSE}
library(sf)
library(readr)
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
library(here)
library(stringr)
library(forcats)
```

load data

```{r}
Standard_lines <- read_tsv(here("data", "Standard_routes_lines.txt"))
Standard_points <- read_tsv(here("data", "Standard_routes_points.txt"))
Standard_geo <- read_sf(here("data", "Standard_routes_geo.gpkg"))
```



### how many routes have been counted each year

```{r}

Standard_points %>% 
  select(yr, karta) %>% 
  distinct() %>% 
  group_by(yr) %>% 
  mutate(n = n()) %>% 
  ggplot(aes(x = yr, y = n))+
  geom_line()+
  geom_point()

```

```{r}

counted_routes <- 
Standard_lines %>% 
  select(yr, karta) %>% 
  distinct() %>% 
  arrange(karta, yr) %>% 
  group_by(karta) %>% 
  mutate(years_counted = 1:n()) %>% 
  right_join(filter(Standard_geo, position == "p1"), .) %>% 
  rename(geometry = geom) %>% 
  ggplot(aes(colour = years_counted))+
  geom_sf(shape =15, size = 0.2)+
  facet_wrap(~yr, ncol = 8)+
  theme_minimal()+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),)+
  scale_colour_viridis_c()

ggsave(here("plots", "counted_routes.pdf"), width = 10, height = 8)

counted_routes

```


### explore rarity (using the last 10 years)

#### average number of observations by year for each species
```{r}
av_bird_abundance_p <-
  Standard_points %>% 
  filter(yr > 2008) %>% 
  group_by(karta) %>% 
  mutate(years_counted = length(unique(yr))) %>% 
  group_by(karta, lat6, englishname) %>% 
  summarise(sum_by_route = sum(abundance), years_counted = unique(years_counted)) %>% 
  mutate(average_abundance = sum_by_route / years_counted) %>% #how often is the species observed on average per year on each route
  group_by(lat6, englishname) %>% 
  summarise(sum_av_abundance = sum(average_abundance)) %>%  #how often is the species observed on average per year in Sweden
  arrange(desc(sum_av_abundance)) %>% 
  ungroup() %>% 
  mutate(rank = 1:n()) 

av_bird_abundance_l <-
  Standard_lines %>% 
  filter(yr > 2008) %>% 
  group_by(karta) %>% 
  mutate(years_counted = length(unique(yr))) %>% 
  group_by(karta, lat6, englishname) %>% 
  summarise(sum_by_route = sum(abundance), years_counted = unique(years_counted)) %>% 
  mutate(average_abundance = sum_by_route / years_counted) %>% #how often is the species observed on average per year on each route
  group_by(lat6, englishname) %>% 
  summarise(sum_av_abundance = sum(average_abundance)) %>%  #how often is the species observed on average per year in Sweden
  arrange(desc(sum_av_abundance)) %>% 
  ungroup() %>% 
  mutate(rank = 1:n()) 

av_bird_abundance <- 
av_bird_abundance_l %>% 
  rename(sum_av_abundance_l = sum_av_abundance) %>% 
  select(-rank) %>% 
  left_join(av_bird_abundance_p) %>% 
  gather(type, abundance, -rank, -lat6, -englishname) %>% 
  mutate(type = case_when(type == "sum_av_abundance_l" ~ "linecounts",
                          type == "sum_av_abundance" ~ "pointcounts"))

spec_rank_plot <- 
  av_bird_abundance %>% 
  ggplot(aes(x= rank, y = abundance, label = englishname))+
  geom_point(aes(colour = type))+
  geom_hline(yintercept = c(1,10), colour = "darkred", linetype = "dashed")+
  geom_text(data = tibble(rank = 10, 
                          abundance = c(1.5,15),
                          label = c("1 observation per year in Sweden",
                                               "10 observations per year in Sweden")),
            aes(x = rank, label = label),
            hjust = "left")+
  scale_y_log10()+
  theme_minimal()+
  labs(y = "average observations of species per year (2009 - 2018)")+
  scale_color_brewer(palette = "Set1")

ggsave(here("plots", "species_ranks.pdf"), height = 6, width = 8)

ggplotly(spec_rank_plot, tooltip = c("englishname"))
```

#### average number and proportion of routes on which each species is observed
```{r}
av_bird_presence_l <- 
 Standard_lines %>% 
  filter(yr > 2008) %>% 
  group_by(karta, lat6, englishname, yr) %>% 
  summarize(presence = sum(abundance)) %>% # sum over all lines
  mutate(presence = as.numeric(presence > 0)) %>% #make abundance p/a
  group_by(yr) %>% 
  mutate(routes_counted = length(unique(karta))) %>% 
  group_by(lat6, englishname, yr) %>% 
  summarise(n_routes = sum(presence),
            routes_counted = unique(routes_counted)) %>% 
  mutate(p_routes = n_routes / routes_counted *100) %>% 
  group_by(lat6, englishname) %>% 
  summarize(n_routes = mean(n_routes),
            p_routes = mean(p_routes)) %>% 
  ungroup() %>% 
  arrange(desc(p_routes)) %>% 
  mutate(rank = 1:n())

av_bird_presence_p <- 
 Standard_points %>% 
  filter(yr > 2008) %>% 
  group_by(karta, lat6, englishname, yr) %>% 
  summarize(presence = sum(abundance)) %>% # sum over all lines
  mutate(presence = as.numeric(presence > 0)) %>% #make abundance p/a
  group_by(yr) %>% 
  mutate(routes_counted = length(unique(karta))) %>% 
  group_by(lat6, englishname, yr) %>% 
  summarise(n_routes = sum(presence),
            routes_counted = unique(routes_counted)) %>% 
  mutate(p_routes = n_routes / routes_counted *100) %>% 
  group_by(lat6, englishname) %>% 
  summarize(n_routes = mean(n_routes),
            p_routes = mean(p_routes)) %>% 
  ungroup() %>% 
  arrange(desc(p_routes)) %>% 
  mutate(rank = 1:n())

av_bird_presence <- 
av_bird_presence_l %>% 
  rename(p_routes_l = p_routes) %>% 
  select(-rank, -n_routes) %>% 
  left_join(av_bird_presence_p) %>% 
  gather(type, proportion, -rank, -lat6, -englishname, -n_routes) %>% 
  mutate(type = case_when(type == "p_routes_l" ~ "linecounts",
                          type == "p_routes" ~ "pointcounts"))

spec_prev_plot <-
  av_bird_presence %>% 
  ggplot(aes(x = rank, y = proportion, label = englishname))+
  geom_point(aes(colour = type))+
  geom_hline(yintercept = c(10,1), colour = "darkred", linetype = "dashed")+
  geom_text(data = tibble(rank = 10, 
                          proportion = c(1.5,0.15),
                          label = c("observed on 10% of the routes on average",
                                    "observed on 1% of the routes on average")),
            aes(x = rank, label = label),
            hjust = "left")+
  scale_y_log10()+
  theme_minimal()+
  labs(y = "average proportion of routes on where \n the species has been observed (2009 - 2018)")+
  scale_color_brewer(palette = "Set1")

ggsave(here("plots", "species_prevalence.pdf"), height = 6, width = 8)

ggplotly(spec_prev_plot, tooltip = c("englishname"))

  
```

#### abundance versus prevalence
```{r}
abun_prop <- 
left_join(select(av_bird_presence, -rank), 
          select(av_bird_abundance, -rank)) %>% 
  filter(type == "linecounts") %>% 
  mutate(proportion_l = signif(proportion, 2)) %>% 
  ggplot(aes(x = abundance, y= proportion, 
             label = englishname, label2 = proportion_l)) + 
  geom_point()+
  scale_x_log10()+
  scale_y_log10()+
  theme_minimal()
  

 ggplotly(abun_prop, tooltip = c("englishname", "proportion_l"))

```

#### detecting flocking species

```{r}
cv_abund <- 
Standard_lines %>% 
  group_by(art, englishname) %>% 
  summarise(n_sites = sum(abundance > 0),
            sum_abund = sum(abundance)) %>%
  filter(n_sites > 100) %>% 
  ungroup() %>% 
  ggplot(aes(x = sum_abund, y = n_sites, label = englishname))+
  geom_point()+
  theme_minimal()+
  scale_x_log10()+
  scale_y_log10()
  

ggplotly(cv_abund, tooltip = c("englishname"))

```


#### rare species 

species that have been observed on average ≤ 1 times in the last 10 years in Sweden

```{r}

av_bird_abundance %>% 
  filter(abundance <=1) %>% 
  right_join(distinct(dplyr::select(Standard_lines, lat6, genus, species)), .)


```

#### overcount on line 8?

Does the observers overcount birds on line 8?

The bird surveillors are asked not to count birds twice between different transects. In theory this leads to that the birders might be careful not to re-count the birds they counted at the end of the previous line, when starting a new line. However, at the end of the route (end of line 8), the birders will re-encounter birds that they might have counted a few hours earlier at the beginning of line 1. Therefore, on could expect to find a higher mean abundance on l8.

```{r}

 Standard_lines %>% 
 filter(art == "247") %>% #Willow warbler
  group_by(karta, position) %>% 
  summarise(abundance = sum(abundance)) %>% 
  {
  ggplot(., aes(x = position, y = abundance))+
  geom_point(position = position_jitter(width = 0.4), alpha = 0.2)+
  stat_summary(geom = "point", fun.y = "median", aes(group = position), colour = "red")+
  geom_hline(yintercept = median(.$abundance))+
  theme_bw()
  }
  
 

```


+ export of bird data summary

```{r}

Bird_summary <- 
av_bird_presence %>% 
  select(-rank) %>% 
  left_join(av_bird_abundance) %>% 
  select(-rank) 

write_csv2(Bird_summary, here("data", "Bird_stats.csv"))
```

### proportion of points counted

```{r}

# 8 routes  have never been counted (because they are in the ocean or very remote / difficult terrain)
routes_not_counted <- unique(Standard_geo$karta)[which(! unique(Standard_geo$karta) %in% Standard_points$karta)]

points_not_counted <- 
Standard_points %>% 
  group_by(karta, position, yr) %>% 
  summarize(abundance = sum(abundance)) %>%
  group_by(karta, position) %>% 
  summarize(abundance = median(abundance)) %>% 
  filter(abundance == 0) %>% 
  select(-abundance) %>% 
  mutate(nevercounted = TRUE) %>% 
  left_join(filter(Standard_geo, type == "point"), .) %>% 
  mutate(nevercounted = ! is.na( nevercounted)) %>% 
  mutate(nevercounted = case_when(karta %in% routes_not_counted ~ TRUE,
                                  TRUE ~ nevercounted))

lines_not_counted <- 
  Standard_lines %>% 
  group_by(karta, position) %>% 
  summarize(prop_counted = mean(prop_counted, na.rm = TRUE)) %>% 
  mutate(prop_counted = round(prop_counted)) %>% 
  left_join(filter(Standard_geo, type == "line"), .) %>% 
  mutate(prop_counted = case_when(karta %in% routes_not_counted ~ 0,
                                  TRUE ~ prop_counted))



write_sf(points_not_counted, here("data", "Points_counted.gpkg"))
write_sf(lines_not_counted, here("data", "Lines_counted.gpkg"))
  
```

