---
title: "preparation of standard route data"
author: "Fabian Roger"
date: "2019-03-01"
output: html_notebook
---

load libraries
```{r, message = FALSE}
library(sf)
library(readr)
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
```

load bird data for Standard routes
```{r}
Standard_data <- read_xlsx("/Volumes/Fabian_Work/bird_routes/public_totalstandard.xlsx")
Standard_geo <- read_xlsx("/Volumes/Fabian_Work/bird_routes/public_standardrutter_koordinater.xlsx")
species_codes <- read_xlsx("/Volumes/Fabian_Work/bird_routes/public_eurolist.xlsx")
```

## reformat geo data

add coordinates for each point and line count

```{r}

# use sweref coordinates and transform to simple feature DF with SWEREF projection
Standard_geo <- 
Standard_geo %>% select(karta, matches("sweref99")) %>% 
  gather(Coord, value, -karta) %>% 
  mutate(position = gsub("(.)_sweref99_.", "\\1", Coord)) %>% 
  mutate(Coord = case_when(grepl("_n", Coord) ~ "LAT",
                           grepl("_o", Coord) ~ "LONG")) %>% 
  spread(Coord, value) %>% 
  st_as_sf(coords = c("LONG", "LAT"), crs = "+proj=utm +zone=33 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs")

# filter out coordinate for lower left corner, dropping centre
Standard_geo_point <- 
  Standard_geo %>% 
  filter(position == "p1") %>% 
  select(-position) 

# expand df to 8 points per route (pointcounts)
Standard_geo_point <- 
  Standard_geo_point %>% 
  left_join(.,
            tibble(karta = rep(Standard_geo_point$karta, each = 8),
                   position = rep(c(paste("p", 1:8, sep = "")), 
                                  length(Standard_geo_point$karta)))) 

# calculate position of each point in reference to starting point. 
# the route always goes first north then east, each side is 2 km.
# there is a point count at each corner and halways for each side

Standard_geo_point <- 
Standard_geo_point %>% 
  mutate(geometry = ifelse(position == "p1", geometry, 
                    ifelse(position == "p2", geometry+c(0,1000),
                    ifelse(position == "p3", geometry + c(0,2000),
                    ifelse(position == "p4", geometry + c(1000,2000),
                    ifelse(position == "p5", geometry + c(2000,2000),
                    ifelse(position == "p6", geometry + c(2000,1000),
                    ifelse(position == "p7", geometry + c(2000,0),
                    geometry + c(1000,0))))))))) %>% 
  mutate(type = "point")

Standard_geo_lines <- 
  Standard_geo_point %>% 
    group_by(karta) %>% 
    slice(c(1,rep(2:n(), each = 2), 1)) %>% 
    mutate(position = rep(paste("l", 1:(n()/2), sep = ""), each = 2)) %>% 
    group_by(karta, position) %>% 
    summarize() %>%
    st_cast("LINESTRING") %>% 
  mutate(type = "line")

Standard_geo <- 
  rbind(Standard_geo_point, Standard_geo_lines) %>% 
  st_set_crs("+proj=utm +zone=33 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs")

```

## plot example

```{r}
Standard_geo %>% 
  filter(karta == "02C2H") %>% 
  mutate(pos = substr(position, 2,2)) %>% 
  ggplot(aes(colour = pos))+
  geom_sf()+
  geom_sf(data = filter(Standard_geo, karta == "02C2H", position == "p1"), 
          colour = "black", shape = 21, size = 4)+
  scale_colour_brewer(palette = "Paired")+
    theme_minimal()

Standard_geo %>% 
  filter(position == "p1") %>% 
  ggplot()+
  geom_sf(shape = 0, size = 0.3)+
  theme_minimal()


```

export geodata

```{r}
write_sf(Standard_geo, here("data", "Standard_routes_geo.gpkg"))
```


### reformating the spatial data

The standard data contain the following information:

+ `persnr`: unique identifier for each observer
+ `karta`: unique identifier for each route
+ `datum`: the date
+ `yr`: the year
+ `art`: the artcode.
    + artcode 000 is used to store the time at which a point has been counted and the time (in minutes) which it has taken to walk a given line
    + artcode 999 is used to say what proportion (1-10, where 1 = 10%) of the line is counted
        + if artcode 999 contains 99 this means no data on the proportion of the line counted is available
        + if artcode 999 contains a 1 in field p1, then mamals have also been counted
+ `p1` - `p8` and `l1 - l8` contain the number of counted birds on the 8 pointcounts and 8 linecounts, respectively. 
+ `pkind` and `lind` are the sum totals for all individuals for that species and route for point and linecounts respectively. 

extracting metadata
```{r}
Standard_data_meta <- 
Standard_data %>% 
  filter(art %in% c("000", "999")) %>% 
  select(-pkind, -lind) %>% 
  gather(position, value, -art, -yr, -datum, -karta, -persnr) %>% 
  spread(art, value) %>% 
  arrange(karta, yr)
```

### point routes

cleaning time for pointcounts
```{r}

# I recode missing values (0 , 99) as NA
# I correct impossible times by assuming that the most likly misstake is a flip in the digits

Standard_data_meta_p <- 
  Standard_data_meta %>% 
    filter(grepl("p", position)) %>% 
    select(-`999`) %>% 
    rename(timepoint = `000`) %>% 
    mutate(timepoint = case_when(timepoint == 9110 ~ 911, #correct data entry error
                                 timepoint == 560 ~ 506, #correct data entry error
                                 timepoint == 973 ~ 937, #correct data entry error
                                 timepoint == 574 ~ 547, #correct data entry error
                                 timepoint == 368 ~ 638, #correct data entry error
                                 timepoint == 9110 ~ 911, #correct data entry error
                                 timepoint == 0 ~ NA_real_, #recode missing value
                                 timepoint == 999 ~ NA_real_, #recode missing value 
                                 TRUE ~ timepoint)) %>% 
    mutate(timepoint = case_when(nchar(as.character(timepoint)) == 3 ~
                                   paste("0", timepoint, sep = ""),
                                 nchar(as.character(timepoint)) == 4 ~
                                   as.character(timepoint))) %>% 
    mutate(timepoint = gsub("(\\d\\d)(\\d\\d)", "\\1:\\2", timepoint)) %>%
    mutate(datum = gsub("(\\d{4})(\\d{2})(\\d{2})", "\\1:\\2:\\3", datum)) 

```


93% of the routes have >= 6 timepoints, < 1 % have none. If we need it we can guess the timpoints for most routes from different years, but we don't do that for now

```{r}

Standard_data_meta_p %>% 
  group_by(karta, datum) %>% 
  mutate(timepoint = is.na(timepoint)) %>% 
  summarise(NA_per_route = sum(timepoint)) %>% 
  group_by(NA_per_route) %>% 
  summarise(n = n()) %>% 
  mutate(prct = round((n / sum(n))*100, 1)) %>% 
  mutate(cum_prct = cumsum(prct))

```


join and export metadata, count data and species names
```{r}

Standard_points <- 
filter(Standard_data, ! art %in% c("000", "999")) %>% # filter out non-species codes
  filter(as.numeric(art) < 701) %>%  # filter out mamals
  select(karta, yr, art, starts_with("p"), -pkind, -persnr) %>% 
  mutate_at(vars(starts_with("p")), function(x) case_when(is.na(x) ~ 0, TRUE ~ x)) %>% 
  gather(position, abundance, -karta, -yr, -art) %>% 
  left_join(select(species_codes, art, englishname, latin, genus, species, lat6)) %>% 
  left_join(Standard_data_meta_p) %>% 
  select(karta, yr, persnr, datum, timepoint, position, art, abundance, englishname, latin, genus, species, lat6) %>% 
  mutate_if(is.numeric, as.integer)

write_tsv(Standard_points, here("data", "Standard_routes_points.txt"))

```


### line routes

cleaning time for pointcounts
```{r}
Standard_data_meta_l <- 
  Standard_data_meta %>% 
    filter(grepl("l", position)) %>% 
    select(-`000`) %>% 
    rename(prop_counted = `999`) %>% 
    mutate(prop_counted = case_when(prop_counted == 99 ~ NA_real_, #recode missing value
                                    TRUE ~ prop_counted)) %>% 
    mutate(datum = gsub("(\\d{4})(\\d{2})(\\d{2})", "\\1:\\2:\\3", datum)) %>% 
  mutate(pos_num = gsub("l(\\d)", "\\1", position)) 


Standard_data_meta_l <- 
  Standard_data_meta_p %>% 
  select(karta, datum, position, timepoint) %>% 
  mutate(pos_num = gsub("p(\\d)", "\\1", position)) %>% 
  select(-position) %>% 
  left_join(Standard_data_meta_l, .) %>% 
  select(-pos_num)

Standard_data_meta_l
```

join and export metadata, count data and species names
```{r}

Standard_lines <- 
filter(Standard_data, ! art %in% c("000", "999")) %>% # filter out non-species codes
  filter(as.numeric(art) < 701) %>%  # filter out mamals
  select(karta, yr, art, starts_with("l"), -lind) %>% 
  mutate_at(vars(starts_with("l")), function(x) case_when(is.na(x) ~ 0, TRUE ~ x)) %>% 
  gather(position, abundance, -karta, -yr, -art) %>% 
  left_join(select(species_codes, art, englishname, latin, genus, species, lat6)) %>% 
  left_join(Standard_data_meta_l) %>% 
  select(karta, yr, persnr, datum, timepoint, position, prop_counted, art,
         abundance, englishname, latin, genus, species, lat6) %>% 
  mutate_if(is.numeric, as.integer)


write_tsv(Standard_lines, here("data", "Standard_routes_lines.txt"))
```




