---
title: "preparation of standard route data"
author: "Fabian Roger"
date: "2019-03-01"
output: html_notebook
---

load libraries
```{r, message = FALSE}
library(sf)
library(readr)
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
```


load bird data for Standard routes
```{r}
Standard_data <- read_xlsx(here("data", "bird_routes", "public_totalstandard.xlsx"))
Standard_geo_lines <- read_sf(here("data", "Standardrutter-shapefiler", "stdruttLinjer.shp"))
Standard_geo_points <- read_sf(here("data", "Standardrutter-shapefiler", "stdruttPunkter.shp"))
species_codes <- read_xlsx(here("data", "bird_routes", "public_eurolist.xlsx"))
```


## reformat geo data
```{r}
Standard_geo_lines <-
Standard_geo_lines %>% 
  st_set_crs(3021) %>% #RT90 projection
  st_transform(3006) %>% #reproject to SWEREF 
  select(KARTA, LINJE, geometry) %>% 
  rename(karta = "KARTA", position = "LINJE") %>% 
  mutate(position = gsub("L", "l", position)) %>% 
  mutate(type = "line")

Standard_geo_points <-
  Standard_geo_points %>% 
  st_set_crs(3021) %>% #RT90 projection
  st_transform(3006) %>% #reproject to SWEREF 
  select(KARTA, PUNK, geometry) %>% 
  rename(karta = "KARTA", position = "PUNK") %>% 
  mutate(position = gsub("P", "p", position)) %>% 
  mutate(type = "point")

Standard_geo <-
  rbind(Standard_geo_lines, Standard_geo_points)
  
```


## plot example

```{r}
Standard_geo %>% 
  filter(karta == "02C2H") %>% 
  mutate(pos = substr(position, 2,2)) %>% 
  ggplot(aes(colour = pos))+
  geom_sf()+
  geom_sf(data = filter(Standard_geo, karta == "02C2H", position == "p1"), 
          colour = "black", shape = 21, size = 4)+
  scale_colour_brewer(palette = "Paired")+
    theme_minimal()

Standard_geo %>% 
  filter(position == "p1") %>% 
  ggplot()+
  geom_sf(shape = 0, size = 0.3)+
  theme_minimal()


```

export geodata

```{r}
write_sf(Standard_geo, here("data", "Standard_routes_geo.gpkg"))
```


### reformating the spatial data

The standard data contain the following information:

+ `persnr`: unique identifier for each observer
+ `karta`: unique identifier for each route
+ `datum`: the date
+ `yr`: the year
+ `art`: the artcode.
    + artcode 000 is used to store the time at which a point has been counted and the time (in minutes) which it has taken to walk a given line
    + artcode 999 is used to say what proportion (1-10, where 1 = 10%) of the line is counted
        + if artcode 999 contains 99 this means no data on the proportion of the line counted is available
        + if artcode 999 contains a 1 in field p1, then mamals have also been counted
+ `p1` - `p8` and `l1 - l8` contain the number of counted birds on the 8 pointcounts and 8 linecounts, respectively. 
+ `pkind` and `lind` are the sum totals for all individuals for that species and route for point and linecounts respectively. 

extracting metadata
```{r}
Standard_data_meta <- 
Standard_data %>% 
  filter(art %in% c("000", "999")) %>% 
  select(-pkind, -lind) %>% 
  gather(position, value, -art, -yr, -datum, -karta, -persnr) %>% 
  spread(art, value) %>% 
  arrange(karta, yr)

pnc <- 
Standard_data_meta %>% 
    filter(grepl("p", position)) %>% 
    select(-`999`) %>% 
    rename(timepoint = `000`) %>% 
  filter(timepoint == 0) 

# Standard_routes_points %>% 
#   filter(paste(karta, yr, position) %in% paste(pnc$karta, pnc$yr, pnc$position)) %>% 
#   group_by(karta, yr, position, timepoint) %>% 
#   summarise(abundance = sum(abundance)) %>% 
#   filter(abundance > 0) %>% 
#   arrange(abundance)
```

### point routes

cleaning time for pointcounts
```{r}

# I recode missing values (0 , 99) as NA or missing
# I correct impossible times by assuming that the most likely misstake is a flip in the digits

Standard_data_meta_p <- 
  Standard_data_meta %>% 
  filter(grepl("p", position)) %>% 
  select(-`999`) %>% 
  rename(timepoint = `000`) %>% 
  mutate(counted = case_when(timepoint == 0 ~ "no",
                              TRUE ~ "yes")) %>% 
  mutate(timepoint = case_when(timepoint == 9110 ~ 911, #correct data entry error
                               timepoint == 560 ~ 506, #correct data entry error
                               timepoint == 973 ~ 937, #correct data entry error
                               timepoint == 574 ~ 547, #correct data entry error
                               timepoint == 368 ~ 638, #correct data entry error
                               timepoint == 9110 ~ 911, #correct data entry error
                               timepoint == 0 ~ NA_real_, #recode missing value indicating that point has not been counted
                               timepoint == 999 ~ NA_real_, #recode missing value 
                               TRUE ~ timepoint)) %>% 
  mutate(timepoint = case_when(nchar(as.character(timepoint)) == 3 ~
                                 paste("0", timepoint, sep = ""),
                               nchar(as.character(timepoint)) == 4 ~
                                 as.character(timepoint))) %>% 
  mutate(timepoint = gsub("(\\d\\d)(\\d\\d)", "\\1:\\2", timepoint)) %>%
  mutate(datum = gsub("(\\d{4})(\\d{2})(\\d{2})", "\\1:\\2:\\3", datum)) 
```


93% of the routes have >= 6 timepoints, < 1 % have none. If we need it we can guess the timepoints for most routes from different years, but we don't do that for now

```{r}

Standard_data_meta_p %>% 
  group_by(karta, datum) %>% 
  mutate(timepoint = is.na(timepoint)) %>% 
  summarise(NA_per_route = sum(timepoint)) %>% 
  group_by(NA_per_route) %>% 
  summarise(n = n()) %>% 
  mutate(prct = round((n / sum(n))*100, 1)) %>% 
  mutate(cum_prct = cumsum(prct))

```


join and export metadata, count data and species names
```{r}

Standard_points <- 
filter(Standard_data, ! art %in% c("000", "999")) %>% # filter out non-species codes
  filter(as.numeric(art) < 701) %>%  # filter out mamals
  select(karta, yr, art, starts_with("p"), -pkind, -persnr) %>% 
  mutate_at(vars(starts_with("p")), function(x) case_when(is.na(x) ~ 0, TRUE ~ x)) %>% 
  gather(position, abundance, -karta, -yr, -art) %>% 
  left_join(select(species_codes, art, englishname, latin, genus, species, lat6)) %>% 
  left_join(Standard_data_meta_p) %>% 
  select(karta, yr, persnr, datum, timepoint, position,counted, art, abundance, englishname, latin, genus, species, lat6) %>% 
  mutate_if(is.numeric, as.integer)

write_tsv(Standard_points, here("data", "Standard_routes_points.txt"))

```


### line routes

cleaning time for line counts
```{r}
Standard_data_meta_l <- 
  Standard_data_meta %>% 
    filter(grepl("l", position)) %>% 
    select(-`000`) %>% 
    rename(prop_counted = `999`) %>% 
    mutate(prop_counted = case_when(prop_counted == 99 ~ NA_real_, #recode missing value
                                    TRUE ~ prop_counted)) %>% 
    mutate(datum = gsub("(\\d{4})(\\d{2})(\\d{2})", "\\1:\\2:\\3", datum)) %>% 
  mutate(pos_num = gsub("l(\\d)", "\\1", position)) 


Standard_data_meta_l <- 
  Standard_data_meta_p %>% 
  select(karta, datum, position, timepoint) %>% 
  mutate(pos_num = gsub("p(\\d)", "\\1", position)) %>% 
  select(-position) %>% 
  left_join(Standard_data_meta_l, .) %>% 
  select(-pos_num)

Standard_data_meta_l
```

## correct species codes
there are some 

```{r}
species_codes %>% 
  filter(art %in% Standard_data$art) %>% 
  select(latin, genus, species) %>% 
  filter(latin != paste(genus, species))
```



join and export metadata, count data and species names
```{r}

Standard_lines <- 
filter(Standard_data, ! art %in% c("000", "999")) %>% # filter out non-species codes
  filter(as.numeric(art) < 701) %>%  # filter out mammals
  select(karta, yr, art, starts_with("l"), -lind) %>% 
  mutate_at(vars(starts_with("l")), function(x) case_when(is.na(x) ~ 0, TRUE ~ x)) %>% 
  gather(position, abundance, -karta, -yr, -art) %>% 
  left_join(select(species_codes, art, englishname, latin, genus, species, lat6)) %>% 
  left_join(Standard_data_meta_l) %>% 
  select(karta, yr, persnr, datum, timepoint, position, prop_counted, art,
         abundance, englishname, latin, genus, species, lat6) %>% 
  mutate_if(is.numeric, as.integer)


write_tsv(Standard_lines, here("data", "Standard_routes_lines.txt"))
```




